---
title: "Bayesian Inference"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/bayesian-inference.html)"
---

## Bayes' Theorem

$$
p(\theta \mid y)
=
\frac{p(y \mid \theta)p(\theta)}{p(y)}
$$

For estimation:

$$
p(\theta \mid y) \propto p(y \mid \theta)p(\theta)
$$

## Pieces

- prior: $p(\theta)$
- likelihood: $p(y \mid \theta)$
- posterior: $p(\theta \mid y)$

The posterior is the updated state of knowledge.

## Normal Mean

```{r}
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

set.seed(2026)
y <- rnorm(20, mean = 32, sd = 4)
sigma <- 4
mu_0 <- 28
tau_0 <- 6

posterior_normal_mean <- function(y, sigma, mu_0, tau_0) {
  n <- length(y)
  tau_n2 <- 1 / (1 / tau_0^2 + n / sigma^2)
  mu_n <- tau_n2 * (mu_0 / tau_0^2 + n * mean(y) / sigma^2)
  list(mu = mu_n, sd = sqrt(tau_n2))
}

post <- posterior_normal_mean(y, sigma, mu_0, tau_0)
post
```

## Prior To Posterior

```{r}
curve_data <- data.frame(mu = seq(20, 40, length.out = 400))
plot_data <- rbind(
  data.frame(type = "prior", mu = curve_data$mu, density = dnorm(curve_data$mu, mu_0, tau_0)),
  data.frame(type = "likelihood", mu = curve_data$mu, density = dnorm(curve_data$mu, mean(y), sigma / sqrt(length(y)))),
  data.frame(type = "posterior", mu = curve_data$mu, density = dnorm(curve_data$mu, post$mu, post$sd))
)

ggplot(plot_data, aes(mu, density, color = type)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c(prior = "#7a6f42", likelihood = "#c46a2b", posterior = "#0c6773")) +
  labs(x = expression(mu), y = "Density")
```

## Sequential Updating

Posterior from experiment 1 becomes prior for experiment 2.

```{r}
set.seed(2026)
y_first <- rnorm(12, mean = 31, sd = sigma)
y_second <- rnorm(12, mean = 33, sd = sigma)

post_first <- posterior_normal_mean(y_first, sigma, mu_0 = 28, tau_0 = 6)
post_second <- posterior_normal_mean(y_second, sigma, mu_0 = post_first$mu, tau_0 = post_first$sd)

data.frame(
  stage = c("prior", "after first", "after second"),
  mean = c(28, post_first$mu, post_second$mu),
  sd = c(6, post_first$sd, post_second$sd)
)
```

## Why brms?

The conjugate example is transparent.

`brms` is for realistic models:

- multilevel models,
- non-Gaussian responses,
- multivariate responses,
- animal models with covariance structures.

## Stan Setup

```bash
Rscript setup/install.R --stan --install-cmdstan
Rscript setup/check.R --stan
```

Use cached artifacts for class when possible.

## brms Pattern

```{r eval=FALSE}
library(brms)

fit <- brm(
  Sepal.Width ~ 1,
  data = iris,
  family = gaussian(),
  prior = set_prior("normal(0, 1)", class = "Intercept"),
  sample_prior = TRUE,
  backend = "cmdstanr"
)
```

## Bayesian Animal Model

$$
\mathbf{y}
=
\mathbf{X}\boldsymbol{\beta}
+ \mathbf{Z}\mathbf{u}
+ \mathbf{e}
$$

$$
\mathbf{u} \sim N(\mathbf{0}, \mathbf{A}\sigma_a^2)
$$

Priors complete the model.

## Take-Home

Bayesian inference is likelihood plus prior information.

The output is a posterior distribution, not only a point estimate.
