---
title: "Bayesian Inference"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/bayesian-inference.html)"
---

## Bayes' Theorem

$$
p(\theta \mid y)
=
\frac{p(y \mid \theta)p(\theta)}{p(y)}
$$

For model fitting:

$$
p(\theta \mid y) \propto p(y \mid \theta)p(\theta)
$$

Likelihood times prior gives the fitted Bayesian model.

## Estimating A Normal Mean

$$
\mathbf{y} \sim N(\mu, \sigma_e^2),
\qquad
\mu \sim N(\mu_\text{prior}, \sigma_\text{prior}^2)
$$

```{r}
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

data(iris)

data.frame(
  trait = "Sepal.Width",
  n = length(iris$Sepal.Width),
  mean = mean(iris$Sepal.Width),
  sd = sd(iris$Sepal.Width)
)
```

## brms Mean Example

This is the live pattern. The rendered site keeps it unevaluated so Stan compilation does not block the build.

```{r eval=FALSE}
library(brms)
library(posterior)

get_prior(Sepal.Width ~ 1, data = iris)

priors <- set_prior("normal(0, 0.2)", class = "Intercept")

fit <- brm(
  Sepal.Width ~ 1,
  data = iris,
  family = gaussian(),
  prior = priors,
  sample_prior = TRUE,
  backend = "cmdstanr"
)
```

## Prior Versus Data Preview

```{r}
set.seed(2026)

prior_samples <- rnorm(4000, mean = 0, sd = 0.2)
estimate_samples <- rnorm(
  4000,
  mean = mean(iris$Sepal.Width),
  sd = sd(iris$Sepal.Width) / sqrt(nrow(iris))
)

samples <- rbind(
  data.frame(type = "prior", samples = prior_samples),
  data.frame(type = "estimate from data", samples = estimate_samples)
)

ggplot(samples, aes(samples, fill = type)) +
  geom_density(alpha = 0.55) +
  scale_fill_manual(values = c("prior" = "#526A83", "estimate from data" = "#68855C")) +
  labs(x = "Mean of Sepal.Width", y = "Density", fill = NULL)
```

## Updating Priors From Existing Evidence

The old lab used a first experiment to inform the prior for a second experiment.

```{r}
update_mean_preview <- function(y, prior_mean, prior_sd) {
  data_sd <- sd(y)
  n <- length(y)
  precision <- 1 / prior_sd^2 + n / data_sd^2
  post_sd <- sqrt(1 / precision)
  post_mean <- (prior_mean / prior_sd^2 + n * mean(y) / data_sd^2) / precision
  c(mean = post_mean, sd = post_sd)
}

iris_first <- iris[1:50, ]
iris_second <- iris[51:nrow(iris), ]

first <- update_mean_preview(iris_first$Sepal.Width, prior_mean = 0, prior_sd = 4)
second <- update_mean_preview(iris_second$Sepal.Width, prior_mean = first["mean"], prior_sd = first["sd"])

data.frame(
  stage = c("wide prior", "after first subset", "after second subset"),
  mean = c(0, unname(first["mean"]), unname(second["mean"])),
  sd = c(4, unname(first["sd"]), unname(second["sd"]))
)
```

## Multivariate Animal Model In brms

$$
\mathbf{Y}
=
\mathbf{X}\mathbf{B}
+ \mathbf{Z}\mathbf{U}
+ \mathbf{E}
$$

```{r eval=FALSE}
brmf <- brmsformula(
  mvbind(milk, fat) ~ 1 +
    factor(lact) +
    herd +
    (1 | g | gr(idA, cov = A)) +
    (1 | p | gr(id)),
  data = D,
  data2 = list(A = A)
)

get_prior(brmf, data = D, data2 = list(A = A), family = gaussian())
make_stancode(brmf, data = D, data2 = list(A = A), family = gaussian())
```

## Cached Artifacts

Expensive Stan fits should be generated ahead of time:

```bash
Rscript scripts/build-model-artifacts.R --stan
```

Then class can load `.rds` artifacts from `artifacts/models/`.

## Resources

- [Stan](https://mc-stan.org/)
- [brms](https://github.com/paul-buerkner/brms)
- [AnimalModels](https://github.com/cheuerde/AnimalModels)
- [BayesianAlphabet](https://github.com/cheuerde/BayesianAlphabet)

## Exercises

1. Change the prior standard deviation in the `brms` mean example.
2. Run the first `brms` model after checking the Stan setup.
3. Use the first 75 iris rows as the first experiment.
4. Inspect `make_stancode()` for the animal-model formula.
