---
title: "Clustering"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/clustering.html)"
---

## Labelled And Unlabelled Views

Clustering learns grouping structure from features without using trusted labels.

```{r}
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

data(iris)
head(iris)
```

For clustering, `Species` is ignored during fitting and used later only as a check.

## Univariate Clustering

```{r}
features <- iris[, c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")]
feature_long <- stack(features)
names(feature_long) <- c("value", "feature")

ggplot(feature_long, aes(value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 35, fill = "#fff8ef", color = "#263238") +
  geom_density(color = "#c46a2b", linewidth = 1) +
  facet_wrap(~feature, scales = "free") +
  labs(x = "Feature value", y = "Density")
```

## Simulate A Two-Component Mixture

$$
p(x)
=
\sum_{k=1}^{K} \lambda_k f_k(x),
\qquad
\sum_{k=1}^{K} \lambda_k = 1
$$

```{r}
set.seed(100)
n <- 1000
theta <- 0.3
y <- ifelse(runif(n) < theta, rnorm(n, 5, 1), rnorm(n, 15, 2))

ggplot(data.frame(y), aes(y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "white", color = "#263238") +
  geom_density(color = "#c46a2b", linewidth = 1) +
  labs(x = "Simulated trait", y = "Density")
```

## A Simple Mixture Fit

Use `kmeans()` first, then estimate normal components.

```{r}
GetProbNormalMixture <- function(y, lambda, mu, sigma) {
  likelihood <- vapply(seq_along(lambda), function(k) {
    lambda[k] * dnorm(y, mean = mu[k], sd = sigma[k])
  }, numeric(length(y)))
  likelihood / rowSums(likelihood)
}

km <- kmeans(matrix(y, ncol = 1), centers = 2, nstart = 25)
cluster <- match(km$cluster, order(as.numeric(km$centers)))
lambda <- as.numeric(tabulate(cluster, nbins = 2)) / length(y)
mu <- as.numeric(tapply(y, cluster, mean))
sigma <- as.numeric(tapply(y, cluster, sd))

data.frame(component = 1:2, lambda = lambda, mu = mu, sigma = sigma)
```

## Component Densities

```{r}
density_grid <- data.frame(y = seq(min(y), max(y), length.out = 400))
density_grid$component_1 <- lambda[1] * dnorm(density_grid$y, mu[1], sigma[1])
density_grid$component_2 <- lambda[2] * dnorm(density_grid$y, mu[2], sigma[2])
density_grid$total <- density_grid$component_1 + density_grid$component_2

ggplot(data.frame(y), aes(y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "white", color = "#263238") +
  geom_line(data = density_grid, aes(y, total), color = "#0c6773", linewidth = 1.1) +
  geom_line(data = density_grid, aes(y, component_1), color = "#c46a2b", linetype = "dashed") +
  geom_line(data = density_grid, aes(y, component_2), color = "#7a6f42", linetype = "dashed") +
  labs(x = "Simulated trait", y = "Density")
```

## Apply To Iris Features

```{r}
petal <- iris$Petal.Length
km_petal <- kmeans(matrix(petal, ncol = 1), centers = 2, nstart = 25)
cluster_petal <- match(km_petal$cluster, order(as.numeric(km_petal$centers)))

iris_cluster <- data.frame(
  Petal.Length = iris$Petal.Length,
  Species = iris$Species,
  cluster = factor(cluster_petal)
)

ggplot(iris_cluster, aes(Petal.Length, fill = cluster)) +
  geom_histogram(aes(y = after_stat(density)), bins = 35, alpha = 0.65, position = "identity") +
  scale_fill_manual(values = c("#0c6773", "#c46a2b")) +
  labs(x = "Petal length", y = "Density", fill = "Cluster")
```

## Check Against Species Labels

```{r}
table(
  cluster = iris_cluster$cluster,
  species = iris_cluster$Species
)
```

Labels are not used for fitting. They are only a sanity check after clustering.

## Optional Stan Version

The previous implementation used Stan for a proper mixture model.

That is statistically cleaner, but not the simplest first implementation.

```{r eval=FALSE}
library(rstan)

fit <- stan(model_code = hcode, data = dat, chains = 1, iter = 2000)
```

## Exercises

1. Change the simulated mixture means.
2. Change `theta` and inspect the fitted weights.
3. Apply the simple mixture fit to another iris feature.
4. Compare the cluster table with `Species`.
