---
title: "Clustering"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/clustering.html)"
---

## Labelled And Unlabelled Views

The `iris` data contain features and labels.

For clustering, ignore the labels during fitting.

```{r}
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

data(iris)
head(iris)
```

## Features Only

```{r}
features <- iris[, c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")]
head(features)
```

## Univariate Clustering

The old lab started by inspecting one feature at a time.

```{r}
feature_long <- stack(features)
names(feature_long) <- c("value", "feature")

ggplot(feature_long, aes(value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 35, fill = "#fff8ef", color = "#263238") +
  geom_density(color = "#c46a2b", linewidth = 1) +
  facet_wrap(~feature, scales = "free") +
  labs(x = "Feature value", y = "Density")
```

## Simulate A Two-Component Mixture

$$
p(x)
=
\sum_{k=1}^{K} \lambda_k f_k(x),
\qquad
\sum_{k=1}^{K} \lambda_k = 1
$$

```{r}
set.seed(100)

n <- 1000
theta <- 0.3
mu1 <- 5
mu2 <- 15
sd1 <- 1
sd2 <- 2

y <- ifelse(
  runif(n) < theta,
  rnorm(n, mean = mu1, sd = sd1),
  rnorm(n, mean = mu2, sd = sd2)
)

simulated <- data.frame(x = seq_along(y), y = y)
```

## Mixture Histogram

```{r}
ggplot(simulated, aes(y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "white", color = "#263238") +
  geom_density(color = "#c46a2b", linewidth = 1) +
  labs(x = "Simulated trait", y = "Density")
```

## A Simple Mixture Fit

The lab uses the same three-step teaching implementation:

1. split with `kmeans()`,
2. estimate $\lambda_k$, $\mu_k$, and $\sigma_k`,
3. compute component probabilities.

## Lab Helper: `GetProbNormalMixture()`

```{r}
GetProbNormalMixture <- function(y, lambda, mu, sigma) {
  likelihood <- vapply(
    seq_along(lambda),
    function(k) lambda[k] * dnorm(y, mean = mu[k], sd = sigma[k]),
    numeric(length(y))
  )

  likelihood / rowSums(likelihood)
}
```

## Lab Helper: `RunMixture()` {.smaller}

```{r}
RunMixture <- function(y, k = 2, seed = 1) {
  set.seed(seed)

  km <- kmeans(matrix(y, ncol = 1), centers = k, nstart = 25)
  center_order <- order(as.numeric(km$centers))
  cluster <- match(km$cluster, center_order)

  lambda <- as.numeric(tabulate(cluster, nbins = k)) / length(y)
  mu <- as.numeric(tapply(y, cluster, mean))
  sigma <- as.numeric(tapply(y, cluster, sd))
  probs <- GetProbNormalMixture(y, lambda = lambda, mu = mu, sigma = sigma)

  dprobs <- as.data.frame(probs)
  names(dprobs) <- paste0("Component_", seq_len(k))
  dprobs$x <- seq_along(y)
  dprobs$y <- y
  dprobs$cluster <- factor(max.col(probs))

  grid <- seq(min(y) - sd(y), max(y) + sd(y), length.out = 400)
  density <- do.call(
    rbind,
    lapply(seq_len(k), function(component) {
      data.frame(
        component = factor(component),
        y = grid,
        density = lambda[component] * dnorm(grid, mu[component], sigma[component])
      )
    })
  )

  list(
    mc = data.frame(component = seq_len(k), lambda = lambda, mu = mu, sigma = sigma),
    dprobs = dprobs,
    density = density
  )
}
```

## Fit Simulated Mixture

```{r}
sim_mixture <- RunMixture(y, k = 2, seed = 242151)
sim_mixture$mc
```

## Component Densities

```{r}
ggplot(simulated, aes(y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "white", color = "#263238") +
  geom_line(data = sim_mixture$density, aes(y, density, color = component), linewidth = 1.1) +
  scale_color_manual(values = c("#0c6773", "#c46a2b")) +
  labs(x = "Simulated trait", y = "Density", color = "Component")
```

## Component Probabilities

```{r}
ggplot(sim_mixture$dprobs, aes(x, y, color = Component_1)) +
  geom_point(alpha = 0.8) +
  scale_color_gradient2(low = "#c46a2b", mid = "#fff8ef", high = "#0c6773", midpoint = 0.5) +
  labs(x = "Observation", y = "Simulated trait", color = "P(component 1)")
```

## Apply To Iris Features

```{r}
seed <- 242151

iris_mixtures <- lapply(features, RunMixture, k = 2, seed = seed)

component_summary <- do.call(
  rbind,
  lapply(names(iris_mixtures), function(feature) {
    out <- iris_mixtures[[feature]]$mc
    out$feature <- feature
    out[, c("feature", "component", "lambda", "mu", "sigma")]
  })
)

component_summary
```

## Check Against Species Labels

```{r}
petal_length <- iris_mixtures[["Petal.Length"]]

iris_cluster <- data.frame(
  Petal.Length = iris$Petal.Length,
  Species = iris$Species,
  cluster = petal_length$dprobs$cluster,
  prob_component_1 = petal_length$dprobs$Component_1
)

table(cluster = iris_cluster$cluster, species = iris_cluster$Species)
```

## Petal-Length Clusters

```{r}
ggplot(iris_cluster, aes(Petal.Length, fill = cluster)) +
  geom_histogram(aes(y = after_stat(density)), bins = 35, alpha = 0.65, position = "identity") +
  scale_fill_manual(values = c("#0c6773", "#c46a2b")) +
  labs(x = "Petal length", y = "Density", fill = "Cluster")
```

## Optional Stan Version

The previous implementation used Stan for a proper mixture model.

```{r eval=FALSE}
library(rstan)

fit <- stan(model_code = hcode, data = dat, chains = 1, iter = 2000)
```

Keep this as an extension after the Bayesian tooling is working.

## Exercises

1. Fit `RunMixture()` to `Sepal.Width`.
2. Change `k = 3` for `Petal.Length`.
3. Simulate two components that overlap more strongly.
4. Compare `Petal.Length` and `Petal.Width`.
