---
title: "Clustering"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/clustering.html)"
---

## Clustering

Clustering learns grouping structure from unlabelled data.

We have features, but no trusted labels.

## Labelled Example

```{r}
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

data(iris)
head(iris)
```

For clustering, `Species` is ignored during fitting.

## Feature Histograms

```{r}
features <- iris[, c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")]
feature_long <- stack(features)
names(feature_long) <- c("value", "feature")

ggplot(feature_long, aes(value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 35, fill = "#fff8ef", color = "#263238") +
  geom_density(color = "#c46a2b", linewidth = 1) +
  facet_wrap(~feature, scales = "free") +
  labs(x = "Feature value", y = "Density")
```

## Mixture Idea

$$
p(x)
=
\sum_{k=1}^{K} \lambda_k f_k(x),
\qquad
\sum_{k=1}^{K} \lambda_k = 1
$$

For a normal mixture:

$$
f_k(x) = N(x \mid \mu_k, \sigma_k^2)
$$

## Simulate A Mixture

```{r}
set.seed(100)
n <- 1000
theta <- 0.3
y <- ifelse(runif(n) < theta, rnorm(n, 5, 1), rnorm(n, 15, 2))

ggplot(data.frame(y), aes(y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "white", color = "#263238") +
  geom_density(color = "#c46a2b", linewidth = 1) +
  labs(x = "Simulated trait", y = "Density")
```

## Simple Implementation

Use `kmeans()` first, then estimate normal components.

```{r}
GetProbNormalMixture <- function(y, lambda, mu, sigma) {
  likelihood <- vapply(seq_along(lambda), function(k) {
    lambda[k] * dnorm(y, mean = mu[k], sd = sigma[k])
  }, numeric(length(y)))
  likelihood / rowSums(likelihood)
}

km <- kmeans(matrix(y, ncol = 1), centers = 2, nstart = 25)
cluster <- match(km$cluster, order(as.numeric(km$centers)))
lambda <- as.numeric(tabulate(cluster, nbins = 2)) / length(y)
mu <- as.numeric(tapply(y, cluster, mean))
sigma <- as.numeric(tapply(y, cluster, sd))

data.frame(component = 1:2, lambda = lambda, mu = mu, sigma = sigma)
```

## Component Probabilities

```{r}
probs <- GetProbNormalMixture(y, lambda, mu, sigma)
dprobs <- data.frame(x = seq_along(y), y = y, prob_one = probs[, 1])

ggplot(dprobs, aes(x, y, color = prob_one)) +
  geom_point(alpha = 0.8) +
  scale_color_gradient2(low = "#c46a2b", mid = "#fff8ef", high = "#0c6773", midpoint = 0.5) +
  labs(x = "Observation", y = "Simulated trait", color = "P(component 1)")
```

## Apply To Iris

```{r}
petal <- iris$Petal.Length
km_petal <- kmeans(matrix(petal, ncol = 1), centers = 2, nstart = 25)
cluster_petal <- match(km_petal$cluster, order(as.numeric(km_petal$centers)))

table(
  cluster = cluster_petal,
  species = iris$Species
)
```

## Take-Home

The old Stan version is useful later.

For the first clustering lab, the simple version is enough to understand the idea.
