---
title: "Clustering"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/clustering.html)"
---

## Clustering

Clustering or *Unsupervised Learning* deals with learning grouping structures from *unlabelled*
data.

Unlabelled data means the absense of trait or phenotypic observations in a breeding context while having *features* available
on a set of individuals.

## Features And Labels

Illustration on a dataset that contains *features* and *labels*.

The features here are `Sepal.Length`,`Sepal.Width`,`Petal.Length` and `Petal.Width`.

The labels for those feature observations are the `Species`.

## Labelled Data

```{r eval=FALSE}
library(pacman)
p_load(ggplot2, knitr, plotly, lubridate, tidyverse, rstan)

# load dataset
data(iris)

# check out features and labels
glimpse(iris)
```

## Unlabelled Data

Now we treat the features as coming from unlabelled individuals, meaning we dont have information on the grouping factor (Species)

```{r eval=FALSE}
# load dataset
data(iris)

# check out features and labels
glimpse(iris %>% select(-Species))
```

## Univariate Clustering

An initial approach is to investigate the indivudal features in isolation from each other and
try to infer a grouping structure.

```{r eval=FALSE}
dplot = pivot_longer(iris, -Species, names_to = "Feature", values_to = "Value")
p = ggplot(dplot, aes(x = Value)) +
  geom_histogram(aes(y = ..density.., fill = Feature), bins = 40) +
  geom_density() +
  facet_wrap(~ Feature)
p
```

## Normal Mixture Model

Some features indicate that the data had been generated not from a single normal process, but
potentially a series of processes.

$$
p(\mathbf{x}) = \sum_{i=1}^{K} \lambda_i f_i(\mathbf{x}), ~~~~~~ with \sum \boldsymbol{\lambda} = 1.
$$

Univariate normal distribution: $f(x) \sim N(\mu, \sigma^2)$.

## Simulate Mixture Data

```{r}
library(ggplot2)
theme_set(theme_minimal(base_size = 16))

# some random data
set.seed(100)
N <- 1000

# mixture probability for the first component
theta <- 0.3
mu1 <- 5
mu2 <- 15
sd1 <- 1
sd2 <- 2
y <- ifelse(runif(N) < theta, rnorm(N, mu1, sd1), rnorm(N, mu2, sd2))

dat = list(
     N = length(y),
     y = y,
     k = 2
)
```

## Raw Mixture

```{r}
dplot = as.data.frame(dat)

ggplot(dplot, aes(x = y)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "white", color = 1) +
  geom_density()
```

## Stan Mixture Program

```{r eval=FALSE}
hcode =
"
data {
    int<lower=1> N;
    int<lower=1> k;
    real y[N];
}
parameters {
    simplex[k] theta;
    real mu[k];
    real<lower=0> tau[k];
}
model {
    real ps[k];
    for (i in 1:k){
        mu[i] ~ normal(0, 1.0e+2);
    }
    for(i in 1:N){
        for(j in 1:k){
            ps[j] <- log(theta[j]) + normal_log(y[i], mu[j], tau[j]);
        }
        increment_log_prob(log_sum_exp(ps));
    }
    tau ~ cauchy(0,5);
}
"
```

## Fit Mixture

```{r eval=FALSE}
hlm = stan(
     model_name = "Gauss mixture",
     model_code = hcode,
     data = dat,
     iter = 2000,
     warmup = 1500,
     verbose = FALSE,
     chains = 1
)

out <- rstan::extract(hlm)

thetas = colMeans(out$theta)
mus = colMeans(out$mu)
sds = colMeans(out$tau)
```

## Mixture Probabilities

```{r eval=FALSE}
GetProbNormalMixture <- function(y, lambda, mu, sigma) {

  lik <- array(as.numeric(NA), dim = c(length(y), length(lambda)))
  for(i in 1:length(lambda)) lik[,i] <- lambda[i] * dnorm(y,mu[i],sigma[i])
  probs <- lik / rowSums(lik)
  return(probs)

}

dnormLambda <- function(x, lambda, mu , sigma) {
  return(lambda * dnorm(x, mu, sigma))
}
```

## Apply To Iris

First, bring everything together into a function that we can call conviently on different data
and different number of mixture components.

```{r eval=FALSE}
RunMixture = function(y, k = 2, niter = 2000, warmup = 1500, seed = NULL) {
  # stan program, fit, posterior extraction, plots
}

seed = 242151

sl_mixture = RunMixture(iris$Sepal.Length, k = 2, seed = seed)
sw_mixture = RunMixture(iris$Sepal.Width, k = 2, seed = seed)
pl_mixture = RunMixture(iris$Petal.Length, k = 2, seed = seed)
pw_mixture = RunMixture(iris$Petal.Width, k = 2, seed = seed)
```

## Inspect Components

```{r eval=FALSE}
kable(sl_mixture$mc)
kable(sw_mixture$mc)
kable(pl_mixture$mc)
kable(pw_mixture$mc)

sl_mixture$p_raw
sw_mixture$p_raw
pl_mixture$p_raw
pw_mixture$p_raw

sl_mixture$p_density
sw_mixture$p_density
pl_mixture$p_density
pw_mixture$p_density
```
