---
title: "Genomic Prediction"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/genomic-prediction.html)"
---

## Goal

Use marker genotypes to predict breeding values.

Two views of the same model:

- marker effects: estimate $\hat{\mathbf{a}}$,
- genomic relationships: estimate $\hat{\mathbf{g}}$ directly.

## Marker Model

$$
\mathbf{y}
=
\mathbf{1}\mu
+ \mathbf{Z}\mathbf{a}
+ \mathbf{e}
$$

$$
\mathbf{a} \sim N(\mathbf{0}, \mathbf{I}\sigma_a^2),
\qquad
\mathbf{e} \sim N(\mathbf{0}, \mathbf{I}\sigma_e^2)
$$

The GEBV is

$$
\hat{\mathbf{g}} = \mathbf{Z}\hat{\mathbf{a}}.
$$

## Simulate Markers

```{r}
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

set.seed(2026)
n_animals <- 160
n_markers <- 300

allele_frequency <- runif(n_markers, 0.05, 0.50)
M <- sapply(allele_frequency, function(p) rbinom(n_animals, 2, p))

center_markers <- function(M, allele_frequency) {
  sweep(M, 2, 2 * allele_frequency, "-")
}

Z <- center_markers(M, allele_frequency)
qtl <- sample(seq_len(n_markers), 12)
a_true <- rep(0, n_markers)
a_true[qtl] <- rnorm(length(qtl), 0, 0.35)
g_true <- as.numeric(Z %*% a_true)
y <- 30 + g_true + rnorm(n_animals, sd = sqrt(var(g_true) * (1 - 0.45) / 0.45))
```

## Phenotype And Genetic Signal

```{r}
ggplot(data.frame(g_true, y), aes(g_true, y)) +
  geom_point(alpha = 0.75, color = "#0c6773") +
  geom_smooth(method = "lm", se = FALSE, color = "#c46a2b") +
  labs(x = "True breeding value", y = "Phenotype")
```

## Ridge Marker BLUP

$$
\hat{\mathbf{a}}
=
\left(\mathbf{Z}'\mathbf{Z} + \lambda\mathbf{I}\right)^{-1}
\mathbf{Z}'(\mathbf{y} - \bar{y}\mathbf{1})
$$

```{r}
fit_ridge_marker <- function(y, M_train, M_new = M_train, lambda = 50) {
  p_train <- colMeans(M_train) / 2
  Z_train <- center_markers(M_train, p_train)
  Z_new <- center_markers(M_new, p_train)
  y_mean <- mean(y)
  marker_effect <- as.numeric(
    solve(crossprod(Z_train) + diag(lambda, ncol(Z_train)), crossprod(Z_train, y - y_mean))
  )
  list(mean = y_mean, marker_effect = marker_effect, prediction = as.numeric(y_mean + Z_new %*% marker_effect))
}

ridge_fit <- fit_ridge_marker(y, M, lambda = 50)
g_ridge <- ridge_fit$prediction - ridge_fit$mean

cor(g_ridge, g_true)
```

## Genomic Relationship Matrix

$$
\mathbf{G}
=
\frac{\mathbf{Z}\mathbf{Z}'}{\operatorname{mean}\left(\operatorname{diag}(\mathbf{Z}\mathbf{Z}')\right)}
$$

```{r}
ZZt <- tcrossprod(Z)
G <- ZZt / mean(diag(ZZt))

g_subset <- as.data.frame(as.table(G[1:35, 1:35]))
names(g_subset) <- c("animal_1", "animal_2", "relationship")

ggplot(g_subset, aes(animal_1, animal_2, fill = relationship)) +
  geom_tile() +
  scale_fill_gradient2(low = "#245c73", mid = "#fff8ef", high = "#c46a2b") +
  coord_equal() +
  labs(x = NULL, y = NULL, fill = "G")
```

## GBLUP

$$
\mathbf{y}
=
\mathbf{1}\mu
+ \mathbf{g}
+ \mathbf{e},
\qquad
\mathbf{g} \sim N(\mathbf{0}, \mathbf{G}\sigma_g^2)
$$

## Equivalence

```{r}
lambda_marker <- 50
lambda_g <- lambda_marker / mean(diag(ZZt))
y_centered <- y - mean(y)

g_gblup <- as.numeric(G %*% solve(G + diag(lambda_g, nrow(G)), y_centered))

data.frame(
  correlation = cor(g_ridge, g_gblup),
  maximum_difference = max(abs(g_ridge - g_gblup))
)
```

## Cross-Validation

```{r}
make_folds <- function(n, k = 5, seed = 1) {
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

cross_validate_ridge <- function(y, M, lambda_values, k = 5, seed = 1) {
  fold <- make_folds(length(y), k = k, seed = seed)
  do.call(rbind, lapply(lambda_values, function(lambda) {
    prediction <- rep(NA_real_, length(y))
    for (f in seq_len(k)) {
      training <- fold != f
      validation <- fold == f
      fit <- fit_ridge_marker(y[training], M[training, , drop = FALSE], M[validation, , drop = FALSE], lambda)
      prediction[validation] <- fit$prediction
    }
    data.frame(lambda = lambda, correlation = cor(prediction, y))
  }))
}

cv_results <- cross_validate_ridge(y, M, c(5, 15, 50, 150, 500), seed = 2026)
cv_results
```

## Validation Curve

```{r}
ggplot(cv_results, aes(lambda, correlation)) +
  geom_line(color = "#0c6773") +
  geom_point(size = 3, color = "#c46a2b") +
  scale_x_log10() +
  labs(x = expression(lambda), y = "Prediction correlation")
```

## Bayesian Alphabet

Ridge:

$$
a_j \sim N(0, \sigma_a^2)
$$

Spike-and-slab:

$$
a_j \mid \gamma_j
\sim
\gamma_j N(0, \sigma_a^2)
+ (1 - \gamma_j)\delta_0
$$

## Take-Home

Genomic prediction is an animal model built from marker similarities.

Marker BLUP and GBLUP are two computational views of the same prediction problem.
