---
title: "Least Squares"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/linear-regression.html)"
---

## A Line Through Data

Choose a line by making vertical distances small.

```{r}
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

set.seed(2026)

milk_data <- data.frame(
  feed_kg = seq(4, 12, length.out = 36)
)

milk_data$milk_kg <- 17.5 + 1.65 * milk_data$feed_kg +
  rnorm(nrow(milk_data), sd = 1.8)
```

## A Candidate Line

```{r}
candidate_beta <- c(intercept = 18, slope = 1.5)

milk_data$candidate_pred <- candidate_beta["intercept"] +
  candidate_beta["slope"] * milk_data$feed_kg

ggplot(milk_data, aes(feed_kg, milk_kg)) +
  geom_segment(
    aes(xend = feed_kg, yend = candidate_pred),
    color = "#b85c38",
    linewidth = 0.5,
    alpha = 0.75
  ) +
  geom_point(size = 2.5, color = "#155f7a") +
  geom_line(aes(y = candidate_pred), color = "#8a4f12", linewidth = 1.1) +
  labs(x = "Concentrate intake (kg/day)", y = "Milk yield (kg/day)")
```

::: notes
The red vertical segments are residuals. Least squares is just a rule for choosing the line with the smallest squared residuals.
:::

## The Least-Squares Problem

$$
\widehat{y}_i = \beta_0 + \beta_1 x_i
$$

$$
\operatorname{SSE}(\boldsymbol{\beta})
=
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \beta_1 x_i\right)^2
$$

$$
\widehat{\boldsymbol{\beta}}
=
\operatorname*{arg\,min}_{\boldsymbol{\beta} \in \mathbb{R}^2}
\operatorname{SSE}(\boldsymbol{\beta})
$$

## `SSE()` In R

```{r}
sse_feed_model <- function(par, data) {
  beta_0 <- par[1]
  beta_1 <- par[2]

  y_hat <- beta_0 + beta_1 * data$feed_kg
  sum((data$milk_kg - y_hat)^2)
}

sse_feed_model(candidate_beta, milk_data)
```

## Grid Search

```{r}
loss_grid <- expand.grid(
  intercept = seq(8, 28, length.out = 90),
  slope = seq(0.4, 2.8, length.out = 90)
)

loss_grid$sse <- mapply(
  FUN = function(intercept, slope) sse_feed_model(c(intercept, slope), milk_data),
  intercept = loss_grid$intercept,
  slope = loss_grid$slope
)

best_grid <- loss_grid[which.min(loss_grid$sse), ]
best_grid
```

## Loss Surface

```{r}
ggplot(loss_grid, aes(intercept, slope, z = sse)) +
  geom_contour_filled(bins = 18) +
  geom_point(
    data = best_grid,
    aes(intercept, slope),
    inherit.aes = FALSE,
    color = "white",
    size = 4
  ) +
  labs(x = "Intercept", y = "Feed response", fill = "SSE")
```

## Solving It With `optim()`

```{r}
fit_optim <- optim(
  par = c(15, 1),
  fn = sse_feed_model,
  data = milk_data,
  method = "BFGS"
)

coef_optim <- setNames(fit_optim$par, c("intercept", "slope"))
coef_optim
```

## The Fitted Line

```{r}
milk_data$optim_pred <- coef_optim["intercept"] + coef_optim["slope"] * milk_data$feed_kg

ggplot(milk_data, aes(feed_kg, milk_kg)) +
  geom_segment(
    aes(xend = feed_kg, yend = optim_pred),
    color = "#b85c38",
    linewidth = 0.45,
    alpha = 0.55
  ) +
  geom_point(size = 2.5, color = "#155f7a") +
  geom_line(aes(y = optim_pred), color = "#8a4f12", linewidth = 1.1) +
  labs(x = "Concentrate intake (kg/day)", y = "Milk yield (kg/day)")
```

## Bound Constraints {.smaller}

$$
\begin{aligned}
\widehat{\boldsymbol{\beta}}
&=
\operatorname*{arg\,min}_{\boldsymbol{\beta} \in \mathbb{R}^2}
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \beta_1 x_i\right)^2 \\
\text{subject to } 0 &\le \beta_1 \le 4
\end{aligned}
$$

```{r}
fit_bounded <- optim(
  par = c(15, 1),
  fn = sse_feed_model,
  data = milk_data,
  method = "L-BFGS-B",
  lower = c(-Inf, 0),
  upper = c(Inf, 4)
)

fit_too_tight <- optim(
  par = c(15, 1),
  fn = sse_feed_model,
  data = milk_data,
  method = "L-BFGS-B",
  lower = c(-Inf, 0),
  upper = c(Inf, 1)
)
```

## Boundary Results

```{r}
data.frame(
  parameter = c("intercept", "slope"),
  unconstrained = coef_optim,
  bounded = fit_bounded$par,
  too_tight = fit_too_tight$par
)
```

If an estimate is exactly on a boundary, the boundary is part of the result.

## Matrix Form

$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{e}
$$

```{r}
y_feed <- milk_data$milk_kg
X_feed <- model.matrix(~ feed_kg, data = milk_data)
head(X_feed)
```

## Closed-Form Solution

$$
\widehat{\boldsymbol{\beta}}
=
\operatorname*{arg\,min}_{\boldsymbol{\beta}}
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})'
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})
$$

$$
\widehat{\boldsymbol{\beta}}
=
(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}
$$

## Same Answer

```{r}
beta_closed <- solve(t(X_feed) %*% X_feed, t(X_feed) %*% y_feed)
coef_lm <- coef(lm(milk_kg ~ feed_kg, data = milk_data))

data.frame(
  method = c("optim", "closed form", "lm"),
  intercept = c(coef_optim["intercept"], beta_closed[1, 1], coef_lm[[1]]),
  slope = c(coef_optim["slope"], beta_closed[2, 1], coef_lm[["feed_kg"]])
)
```

## Intercept-Only

```{r}
X_mean <- matrix(1, nrow = length(y_feed), ncol = 1)

solve(t(X_mean) %*% X_mean, t(X_mean) %*% y_feed)
mean(y_feed)
```

The intercept-only least-squares estimate is the sample mean.

## More Design Matrices

```{r}
data(airquality)
dat <- na.omit(airquality[, c("Wind", "Temp", "Month")])
y <- dat$Wind
X_temp <- model.matrix(~ Temp, data = dat)
head(X_temp)
```

## Multiple Regression

```{r}
X_multi <- model.matrix(~ Temp + Month, data = dat)
head(X_multi)
```

Design matrices can contain continuous predictors, factors, interactions, and transformations.

## Ridge Regression

$$
\widehat{\boldsymbol{\beta}}_{\lambda}
=
\operatorname*{arg\,min}_{\boldsymbol{\beta}}
\left[
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})'
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})
+ \lambda \boldsymbol{\beta}'\boldsymbol{\beta}
\right]
$$

## Why Ridge Matters Here

Ridge introduces shrinkage.

Shrinkage is the core idea behind:

- random effects,
- BLUP,
- genomic prediction,
- Bayesian priors.

## Bridge To Animal Models {.smaller}

$$
\mathbf{y}
=
\mathbf{X}\boldsymbol{\beta}
+ \mathbf{Z}\mathbf{u}
+ \mathbf{e}
$$

$$
\mathbf{u} \sim N(\mathbf{0}, \mathbf{G}), \qquad
\mathbf{e} \sim N(\mathbf{0}, \mathbf{R})
$$

Known variance components turn BLUP into penalized least squares.

## Take-Home

Least squares is optimization first.

The closed-form solution is a shortcut for this specific objective.

The design matrix is the bridge from simple lines to animal models.

## Exercises

1. Change the candidate intercept and slope.
2. Make the upper bound on the slope too small.
3. Compare the closed-form solution with `lm()`.
4. Add another predictor to the design matrix.
