---
title: "Linear Regression"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/linear-regression.html)"
---

```{r}
#| include: false
library(ggplot2)

residual_square_data <- function(data, x_name, y_name, pred_name, every = 12, scale = 0.35) {
  keep <- seq(1, nrow(data), by = every)
  out <- data[keep, , drop = FALSE]
  residual <- out[[y_name]] - out[[pred_name]]

  x_range <- diff(range(data[[x_name]], na.rm = TRUE))
  y_range <- diff(range(c(data[[y_name]], data[[pred_name]]), na.rm = TRUE))

  if (!is.finite(x_range) || x_range == 0) x_range <- 1
  if (!is.finite(y_range) || y_range == 0) y_range <- 1

  side_x <- abs(residual) * (x_range / y_range) * scale

  out$xmin <- out[[x_name]]
  out$xmax <- out[[x_name]] + side_x
  out$ymin <- pmin(out[[y_name]], out[[pred_name]])
  out$ymax <- pmax(out[[y_name]], out[[pred_name]])
  out
}
```

## Linear Regression

Linear Regression deals with fitting predicted values based on input features
and regression weights to observations.

In ordinary least squares, we aim to find regression weights that minimize
the mean squared error between predicted and observed values:

$$
\hat{\beta} = arg min \sum_{i = 1}^{n}  (y_i - x_i \beta_i)^2
$$

## Least-Squares Geometry

$$
e_i = y_i - \hat{y}_i
$$

$$
\operatorname{MSE} = \frac{1}{n}\sum_{i=1}^{n} e_i^2
$$

The full regression plots show residuals as vertical segments.

The toy plot isolates the squared-error pieces.

## Toy Least-Squares Piece

```{r}
toy_dat = data.frame(
  x = 1:10,
  y = c(2.1, 2.9, 2.7, 4.0, 4.8, 5.4, 5.1, 6.7, 7.4, 7.9)
)

toy_fit = lm(y ~ x, data = toy_dat)
toy_dat$pred = predict(toy_fit)

toy_squares = residual_square_data(toy_dat, "x", "y", "pred", every = 1, scale = 0.55)

ggplot(toy_dat, aes(x = x, y = y)) +
  geom_rect(
    data = toy_squares,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill = "#c46a2b",
    color = "#c46a2b",
    alpha = 0.18
  ) +
  geom_segment(aes(xend = x, yend = pred), color = "#155f7a", alpha = 0.65) +
  geom_point(size = 2.2) +
  geom_line(aes(y = pred), color = "red", linewidth = 1) +
  labs(x = "x", y = "y")
```

## Feature / Design Matrix

A Feature or Design Matrix maps observations to individual feature criteria.

The most fundamental feature matrix is a column vector with all 1s, meaning every observation
has the same expression for that feature.

In that case, the feature serves as a mapping to the overall
mean of the observations.

## Estimate The Mean With OLS

```{r}
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

# load example data
data(airquality)

# our response vector is Wind
y = airquality$Wind

# create a design matrix for the mean
X = matrix(data = 1, nrow = length(y), ncol = 1)
colnames(X) = "Intercept"

str(X)
```

## MSE Function

```{r}
mse <- function(par, y, X) {

  # the weights are stored in "par", the length of that vector depends
  # on the number of columns in X (number of features)

  # predicted value
  pred = X %*% par

  # mean squared error
  mse = mean((y - pred)**2)

  return(mse)

}
```

## Fit The Intercept

```{r}
p <- rep(0, ncol(X))
names(p) <- colnames(X)

res <- optim(
       par = p,
       fn = mse,
       y = y,
       X = X
            )

# our solution
res$par
```

## Predicted vs Observed

```{r}
pred = X %*% res$par

dat = data.frame(
     y = y,
     pred = as.numeric(pred),
     id = seq_along(y)
     )

ggplot(dat, aes(x = y, y = pred)) +
 geom_point() +
 geom_abline(intercept = 0, slope = 1, color = "red") +
 labs(x = "Observed Wind", y = "Predicted Wind")
```

## Raw Data And Intercept

```{r}
ggplot(dat, aes(x = id, y = y)) +
  geom_segment(aes(xend = id, yend = pred), color = "#155f7a", alpha = 0.45) +
  geom_point() +
  geom_hline(yintercept = res$par, color = "red") +
  labs(x = "Observation", y = "Wind")
```

## Multiple Linear Regression

When there is more than one feature in the design matrix we call that Multiple Linear Regression.

The overall principle of the problem remains the same: We aim to minimize the mean squared error
of our predictions.

## Add Temperature

```{r}
# our response vector is Wind
y = airquality$Wind

# now we also want another feature: Temp
temp = airquality$Temp

# create a design matrix for the mean
X = matrix(data = 1, nrow = length(y), ncol = 1)

# add temperature as a feature to our design matrix
X = cbind(
   X,
   temp
   )

colnames(X) = c("Intercept", "Temp")
```

## Wind vs Temp

```{r}
ggplot(airquality, aes(x = Temp, y = Wind)) +
  geom_point() +
  labs(x = "Temp", y = "Wind")
```

## Fit Multiple Regression

```{r}
p <- rep(0, ncol(X))
names(p) <- colnames(X)

res <- optim(
       par = p,
       fn = mse,
       y = y,
       X = X
            )

# our solution
res$par
```

## Fitted Slope

```{r}
pred = X %*% res$par

dat = data.frame(
     Wind = y,
     Temp = airquality$Temp,
     pred = as.numeric(pred),
     id = 1:length(y)
     )

ggplot(dat, aes(x = Temp, y = Wind)) +
  geom_segment(aes(xend = Temp, yend = pred), color = "#155f7a", alpha = 0.45) +
  geom_point() +
  geom_abline(intercept = res$par[1], slope = res$par[2], color = "red") +
  labs(x = "Temp", y = "Wind")
```

## Ridge Regression

$$
\min_{\beta} \left( \sum_{i=1}^{n} \left( y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij} \right)^2 + \lambda \sum_{j=1}^{p} \beta_j^2 \right)
$$

## Lambda Solution Space {.smaller}

```{r}
library(BGLR)
data(wheat)

subset = 1:100

y_ridge = wheat.Y[subset,2]
M_ridge = wheat.X[subset,1:300]

M_centered = scale(M_ridge, center = TRUE, scale = FALSE)
y_centered = y_ridge - mean(y_ridge)
n_obs = length(y_ridge)

fit_ridge_lambda <- function(lambda, y_centered, M_centered) {
  u = solve(
    crossprod(M_centered) / n_obs + diag(lambda, ncol(M_centered)),
    crossprod(M_centered, y_centered) / n_obs
  )

  pred = mean(y_ridge) + M_centered %*% u
  mse = mean((y_ridge - pred)**2)
  penalty = lambda * sum(u**2)

  data.frame(lambda = lambda, mse = mse, penalty = penalty,
             criteria = mse + penalty, sd_marker_effect = sd(as.numeric(u)))
}

lambda_grid = 10^seq(-5, 1, length.out = 80)
ridge_path = do.call(rbind, lapply(lambda_grid, fit_ridge_lambda,
                                   y_centered = y_centered, M_centered = M_centered))
best_lambda = ridge_path[which.min(ridge_path$criteria), ]
best_lambda
```

## Ridge Criterion

```{r}
ggplot(ridge_path, aes(x = log10(lambda), y = criteria)) +
  geom_line(color = "#155f7a", linewidth = 1) +
  geom_point(data = best_lambda, color = "#c46a2b", size = 3) +
  labs(x = "log10(lambda)", y = "Ridge criterion")
```

## Marker Shrinkage

```{r}
ggplot(ridge_path, aes(x = log10(lambda), y = sd_marker_effect)) +
  geom_line(color = "#155f7a", linewidth = 1) +
  geom_point(data = best_lambda, color = "#c46a2b", size = 3) +
  labs(x = "log10(lambda)", y = "SD of marker effects")
```

## Ridge In The Original Lab

```{r eval=FALSE}
library(BGLR)
data(wheat)

subset = 1:100

y = wheat.Y[subset,2]
M = wheat.X
M = M[subset,1:300]

X = matrix(data = 1, nrow = length(y), ncol = 1)
colnames(X) = "intercept"
```

## Ridge Objective

```{r eval=FALSE}
mse_ridge <- function(par, y, M, X) {

  b = par[1:ncol(X)]
  mu = X %*% b

  u = par[(ncol(X) + 1):(length(par) - 1)]
  pred = mu + M %*% u

  mse = mean((y - pred)**2)
  ridge_param = par[length(par)]
  ridge = ridge_param * sum(u**2)

  criteria = mse + ridge

  if(ridge_param < 0) {
    criteria = 9999999
  }

  return(criteria)
}
```

## Mixed Model Comparison

Now we compare the results from the optimization to an equivalent mixed model approach with a single randome effect:

$$\begin{bmatrix} X'X & X'Z \\ Z'X & Z'Z+I\lambda \end{bmatrix} \begin{bmatrix} \beta \\ u \end{bmatrix} = \begin{bmatrix} X'y \\ Z'y \end{bmatrix}$$

```{r eval=FALSE}
library(rrBLUP)
mod_rr = mixed.solve(y = y, X = X, Z = M)

ridge_mme = mod_rr$Ve / mod_rr$Vu
pred_mme = X %*% mod_rr$beta + M %*% mod_rr$u
```

## Advanced Models

[Mixed Model](https://gist.github.com/cheuerde/24b8c7e8010d34b8212b)
