---
title: "Maximum Likelihood"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/maximum-likelihood.html)"
---

## Definition

Optimization asked:

$$
\widehat{\theta} = \operatorname*{arg\,min}_{\theta} f(\theta)
$$

Maximum likelihood chooses $f(\theta)$ from a probability model.

## Likelihood And Log-Likelihood

For independent observations:

$$
L(\theta; \mathbf{y})
=
\prod_{i=1}^{n} p(y_i \mid \theta)
$$

$$
\log L(\theta; \mathbf{y})
=
\sum_{i=1}^{n} \log p(y_i \mid \theta)
$$

## In R

`optim()` minimizes.

So we return the negative log-likelihood:

$$
-\sum_{i=1}^{n} \log p(y_i \mid \theta)
$$

## Normal Distribution {.code-tight}

$$
f(y) = \frac{1}{\sigma \sqrt{2\pi}}
\exp\left[-\frac{1}{2}\left(\frac{y - \mu}{\sigma}\right)^2\right]
$$

```{r}
#| fig-show: hide
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

curve(
  dnorm(
    x,
    mean = 10,
    sd = 2
  ),
  from = 0,
  to = 20,
  main = "Normal distribution"
)
```

## Normal Density Plot

```{r}
#| echo: false
curve(
  dnorm(
    x,
    mean = 10,
    sd = 2
  ),
  from = 0,
  to = 20,
  main = "Normal distribution"
)
```

## Finding The Mean

```{r}
set.seed(2026)

mean = 4.5
sd = 1
n = 1000

y = rnorm(n, mean, sd)

lik <- function(par, y, sd = 1) {
  mean = par[1]
  L <- dnorm(x = y, mean = mean, sd = sd, log = TRUE)
  return(-sum(L))
}
```

## Estimate Mean

```{r}
p <- c(0)
names(p) <- c("mu")

ML <- optim(p, lik, y = y)

data.frame(
  method = c("Maximum Likelihood", "Empirical"),
  mu = c(ML$par, mean(y))
)
```

## Finding Mean And Standard Deviation

```{r}
sdlik <- function(x) {
  res <- 0
  if (x > 0) res = 1
  return(log(res))
}

lik <- function(par, y) {
  mean = par[1]
  sd = par[2]

  L2 <- sdlik(sd)
  if (!is.finite(L2)) return(Inf)

  L1 <- dnorm(x = y, mean = mean, sd = sd, log = TRUE)
  return(-sum(L1 + L2))
}
```

## Estimate Both

```{r}
p <- c(0, 1)
names(p) <- c("mu", "sigma")

ML <- optim(p, lik, y = y, method = "Nelder-Mead")

rbind(
  ML = ML$par,
  empirical = c(mean(y), sd(y))
)
```

## Estimating Correlation {.smaller}

```{r}
library(mvtnorm)

corlik <- function(x) {
  res <- 0
  if (x >= -1 & x <= 1) res = 1
  return(log(res))
}

varlik <- function(x) {
  res <- 0
  if (x > 0) res = 1
  return(log(res))
}
```

## Build Sigma

```{r}
lik <- function(par, y) {
  L2 <- corlik(par[5])
  L3 <- varlik(par[3])
  L4 <- varlik(par[4])
  if (!all(is.finite(c(L2, L3, L4)))) return(Inf)

  covAB <- par[5] * (sqrt(par[3]) * sqrt(par[4]))
  sigma <- array(c(par[3], covAB, covAB, par[4]), dim = c(2, 2))

  L1 <- tryCatch(
    dmvnorm(x = y, mean = c(par[1], par[2]), sigma = sigma, log = TRUE),
    error = function(e) -Inf
  )
  if (!all(is.finite(L1))) return(Inf)

  return(-sum(L1 + L2 + L3 + L4))
}
```

## Estimate Correlation

```{r}
methods = c("Nelder-Mead", "BFGS", "CG", "L-BFGS-B", "SANN", "Brent")

p <- c(0, 0, 1, 1, 0.2)
names(p) <- c("mu1", "mu2", "tauA", "tauB", "rho")

set.seed(2026)
n = 10
tauA <- 3
tauB <- 10
rho = 0.7
covAB <- rho * (sqrt(tauA) * sqrt(tauB))
muA <- 5
muB <- 20
sigma <- array(c(tauA, covAB, covAB, tauB), dim = c(2, 2))
Y <- rmvnorm(n, c(muA, muB), sigma)

ML <- optim(p, lik, y = Y, method = methods[1])

res <- as.list(ML$par)
res$covAB <- res$rho * (sqrt(res$tauA) * sqrt(res$tauB))
```

## Correlation Results

```{r}
out <- rbind(
  unlist(res),
  c(colMeans(Y), var(Y[, 1]), var(Y[, 2]), cor(Y)[1, 2], cov(Y)[1, 2])
)
colnames(out) <- names(res)
rownames(out) <- c("ML", "empirical")

out
```

## Animal Model Likelihood

$$
\mathbf{y}
=
\mathbf{X}\mathbf{b}
+ \mathbf{a}
+ \mathbf{e}
$$

$$
\mathbf{y} \sim MVN(\mathbf{X}\mathbf{b},
\mathbf{A}\sigma_a^2 + \mathbf{I}\sigma_e^2)
$$

## Animal Likelihood {.smaller}

```{r}
set.seed(2026)
n_animals <- 20
A <- diag(n_animals)
A[1, 2] <- A[2, 1] <- 0.5
A[3, 4] <- A[4, 3] <- 0.5
A[5, 6] <- A[6, 5] <- 0.25

X <- matrix(1, n_animals, 1)
b <- 30
var_a <- 4
var_e <- 9

a <- as.numeric(rmvnorm(1, mean = rep(0, n_animals), sigma = A * var_a))
y <- as.numeric(X %*% b + a + rnorm(n_animals, sd = sqrt(var_e)))
```

## Animal Likelihood Function {.smaller}

```{r}
lik <- function(par, y, X, A) {
  L2 <- varlik(par[1])
  L3 <- varlik(par[2])
  if (!all(is.finite(c(L2, L3)))) return(Inf)

  b <- matrix(c(par[3:(2 + ncol(X))]), ncol = 1)
  G <- A * par[2]
  R <- diag(length(y)) * par[1]
  V <- G + R
  mu <- X %*% b

  L1 <- tryCatch(
    dmvnorm(x = y, mean = as.numeric(mu), sigma = V, log = TRUE),
    error = function(e) -Inf
  )
  if (!all(is.finite(L1))) return(Inf)

  return(-sum(L1 + L2 + L3))
}
```

## Estimate Animal Model

```{r}
getp <- function(X, b = 0.1, var_e = 1, var_a = 1) {
  p <- c(var_e, var_a, rep(b, ncol(X)))
  names(p) <- c("var_e", "var_a", paste("b", 0:(ncol(X) - 1), sep = ""))
  p
}

ML <- optim(getp(X, b = mean(y)), lik, X = X, A = A, y = y, method = methods[1])

rbind(
  ML = ML$par,
  true = c(var_e, var_a, b)
)
```

## Take-Home

Maximum likelihood is not a new algorithm.

It is an optimization problem where the objective comes from a probability model.

The code pattern is:

1. collect unknowns in `p`,
2. build the likelihood in `lik()`,
3. run `optim(p, lik, ...)`.

## Exercises

1. Change starting values in `p`.
2. Estimate `mu` and `sigma` with `n = 20`.
3. Change the true value of `rho`.
4. Change the relationship matrix `A`.
