---
title: "Multivariate Clustering"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/multivariate-clustering.html)"
---

## Multivariate Clustering

Use several features at once.

Animal breeding example:

- marker genotypes,
- population structure,
- family or line separation,
- outlier detection.

## Mixture Model

For $\mathbf{x}_i \in \mathbb{R}^p$:

$$
p(\mathbf{x}_i)
=
\sum_{k=1}^K \pi_k
N_p(\mathbf{x}_i \mid \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k)
$$

## Ellipse Geometry

$$
d_{ik}^2
=
(\mathbf{x}_i - \boldsymbol{\mu}_k)'
\boldsymbol{\Sigma}_k^{-1}
(\mathbf{x}_i - \boldsymbol{\mu}_k)
$$

For two features, constant $d_{ik}^2$ gives an ellipse.

## Simulate Lines

```{r}
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 16))

simulate_structured_genotypes <- function(n_per_line = c(70, 70, 70), n_markers = 450, seed = 2026) {
  set.seed(seed)
  base_frequency <- runif(n_markers, 0.15, 0.45)
  line_shift <- c(-0.06, 0.00, 0.07)
  marker_list <- lapply(seq_along(n_per_line), function(line) {
    p <- base_frequency + line_shift[line] + rnorm(n_markers, 0, 0.015)
    p <- pmin(pmax(p, 0.02), 0.98)
    sapply(p, function(marker_p) rbinom(n_per_line[line], 2, marker_p))
  })
  list(M = do.call(rbind, marker_list),
       line = factor(rep(paste("line", LETTERS[seq_along(n_per_line)]), n_per_line)))
}

geno <- simulate_structured_genotypes()
```

## PCA

$$
\mathbf{X}
=
\mathbf{U}\mathbf{D}\mathbf{V}'
$$

The scores $\mathbf{U}\mathbf{D}$ summarize marker variation.

```{r}
pca <- prcomp(geno$M, center = TRUE, scale. = TRUE)
pc_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  PC3 = pca$x[, 3],
  line = geno$line
)

ggplot(pc_data, aes(PC1, PC2, color = line)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("#0c6773", "#c46a2b", "#7a6f42")) +
  labs(x = "PC1", y = "PC2", color = "Known line")
```

## K-Means

$$
\sum_{k=1}^K \sum_{i:z_i=k}
\left\lVert \mathbf{x}_i - \boldsymbol{\mu}_k \right\rVert^2
$$

```{r}
set.seed(2026)
km <- kmeans(pc_data[, c("PC1", "PC2", "PC3")], centers = 3, nstart = 50)
pc_data$kmeans_cluster <- factor(km$cluster)

table(cluster = pc_data$kmeans_cluster, line = pc_data$line)
```

## K-Means Plot

```{r}
ggplot(pc_data, aes(PC1, PC2, color = kmeans_cluster)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("#0c6773", "#c46a2b", "#7a6f42")) +
  labs(x = "PC1", y = "PC2", color = "K-means")
```

## Mixture Helpers

```{r include=FALSE}
dmvn_density <- function(X, mean, Sigma) {
  X <- as.matrix(X)
  p <- ncol(X)
  Sigma <- Sigma + diag(1e-6, p)
  centered <- sweep(X, 2, mean, "-")
  Sigma_inv <- solve(Sigma)
  quadratic <- rowSums((centered %*% Sigma_inv) * centered)
  log_det <- as.numeric(determinant(Sigma, logarithm = TRUE)$modulus)
  exp(-0.5 * (p * log(2 * pi) + log_det + quadratic))
}

fit_mvn_mixture <- function(X, k = 3, max_iter = 150, tolerance = 1e-6) {
  X <- as.matrix(X)
  n <- nrow(X)
  p <- ncol(X)
  km_init <- kmeans(X, centers = k, nstart = 30)
  pi <- as.numeric(table(factor(km_init$cluster, levels = seq_len(k)))) / n
  mu <- do.call(rbind, lapply(seq_len(k), function(j) colMeans(X[km_init$cluster == j, , drop = FALSE])))
  Sigma <- lapply(seq_len(k), function(j) cov(X[km_init$cluster == j, , drop = FALSE]) + diag(1e-6, p))
  loglik <- rep(NA_real_, max_iter)
  for (iteration in seq_len(max_iter)) {
    density <- vapply(seq_len(k), function(j) pi[j] * dmvn_density(X, mu[j, ], Sigma[[j]]), numeric(n))
    responsibility <- density / rowSums(density)
    nk <- colSums(responsibility)
    pi <- nk / n
    for (j in seq_len(k)) {
      mu[j, ] <- colSums(X * responsibility[, j]) / nk[j]
      centered <- sweep(X, 2, mu[j, ], "-")
      Sigma[[j]] <- crossprod(centered * sqrt(responsibility[, j])) / nk[j] + diag(1e-6, p)
    }
    density_updated <- vapply(seq_len(k), function(j) pi[j] * dmvn_density(X, mu[j, ], Sigma[[j]]), numeric(n))
    loglik[iteration] <- sum(log(rowSums(density_updated)))
    if (iteration > 1 && abs(loglik[iteration] - loglik[iteration - 1]) < tolerance) break
  }
  density <- vapply(seq_len(k), function(j) pi[j] * dmvn_density(X, mu[j, ], Sigma[[j]]), numeric(n))
  responsibility <- density / rowSums(density)
  list(pi = pi, mu = mu, Sigma = Sigma, responsibility = responsibility,
       cluster = max.col(responsibility), loglik = loglik[max(which(!is.na(loglik)))])
}

ellipse_points <- function(center, Sigma, level = 0.90, n = 200) {
  angle <- seq(0, 2 * pi, length.out = n)
  circle <- cbind(cos(angle), sin(angle)) * sqrt(qchisq(level, df = 2))
  points <- circle %*% chol(Sigma + diag(1e-6, 2))
  data.frame(PC1 = points[, 1] + center[1], PC2 = points[, 2] + center[2])
}
```

## MVN Mixture

```{r}
X2 <- as.matrix(pc_data[, c("PC1", "PC2")])
set.seed(2026)
mvn_fit <- fit_mvn_mixture(X2, k = 3)

pc_data$mvn_cluster <- factor(mvn_fit$cluster)
pc_data$mvn_uncertainty <- 1 - apply(mvn_fit$responsibility, 1, max)

table(cluster = pc_data$mvn_cluster, line = pc_data$line)
```

## Mixture Ellipses

```{r}
ellipse_data <- do.call(rbind, lapply(seq_len(3), function(j) {
  out <- ellipse_points(mvn_fit$mu[j, ], mvn_fit$Sigma[[j]], level = 0.90)
  out$cluster <- factor(j)
  out
}))

ggplot(pc_data, aes(PC1, PC2, color = mvn_cluster)) +
  geom_point(aes(alpha = 1 - mvn_uncertainty)) +
  geom_path(data = ellipse_data, aes(PC1, PC2, color = cluster), linewidth = 1) +
  scale_color_manual(values = c("#0c6773", "#c46a2b", "#7a6f42")) +
  scale_alpha(range = c(0.35, 0.9), guide = "none") +
  labs(x = "PC1", y = "PC2", color = "Mixture")
```

## Choosing K

$$
\operatorname{BIC}
=
-2\ell(\hat{\theta}) + q\log(n)
$$

$$
q = (K - 1) + Kp + K\frac{p(p + 1)}{2}
$$

## Take-Home

PCA helps expose marker-based population structure.

K-means gives hard clusters; mixture models give probabilities and uncertainty.
