---
title: "Introduction to R"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/r-intro.html)"
---

## Get R

[From CRAN](https://cran.r-project.org/)

[RStudio](https://posit.co/download/rstudio-desktop/)

Ubuntu:

```bash
sudo apt install r-base
```

## Some Introduction Material

From the developers: [Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf)

RStudio: [Documentation](https://docs.rstudio.com/)

[R for Data Science](https://r4ds.had.co.nz/introduction.html)

## Background

- R is a general programming language with a specialization for statistics.
- R is generally used through a REPL and interpreted live.
- Functional programming paradigm: all about objects and functions.
- Major appeal: packages, graphics, and scientific workflows.
- Major drawback: scripts can grow without structure.

## Assigning And Manipulating Variables

Note: R has two assignment operators, `<-` and `=`.

```{r}
x = 5
print(x)

x

y = x
print(y)

x = 3

print(x)
print(y)
```

## Working With Data Frames

```{r}
dat = data.frame(
  name = c("Adam", "Paul", "Elly"),
  age = c(22, 43, 54)
)

str(dat)

names = dat$name
ages = dat[, "age"]

str(names)
str(ages)
```

## Subsetting Data Frames

```{r}
dat[dat$name == "Paul", ]
dat[dat$age %in% c(22, 54), ]

dat$id = 1:nrow(dat)
dat[, "id_2"] = dat$id + 2

print(dat)

dat$id[dat$name == "Adam"] <- 12
print(dat)
```

## Reading And Writing Data

```{r}
dat = data.frame(
  name = c("Adam", "Paul", "Elly"),
  age = c(22, 43, 54)
)

write.table(dat, file = "my_data.csv", row.names = FALSE, sep = ",")

dat_read = read.table(file = "my_data.csv", sep = ",", header = TRUE)
str(dat_read)
```

## Vectors And Matrices

```{r}
x = as.integer(1)
str(x)
is.vector(x)

y = rnorm(10)
str(y)

D = matrix(data = rnorm(5 * 5), nrow = 5, ncol = 5)
D
class(D)
```

## Matrix Helpers

```{r}
means = colMeans(D)
means

d = diag(D)
d
```

## Data Exploration And Visualization

```{r}
library(dlookr)
library(ggplot2)

data(iris)
str(iris)

dlookr::diagnose(iris)
```

## Scatterplot

```{r}
p = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

p
```

## Histogram

```{r}
p2 = ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram(aes(fill = Species), bins = 20) +
  facet_wrap(~ Species)

p2
```

## Boxplot

```{r}
p3 = ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot()

p3
```

## Writing Functions

All operators in R are functions.

```{r eval=FALSE}
?sum
?`+`
fix(`+`)
```

## A Mean Function

```{r}
my_mean_function = function(y) {

  n = length(y)
  sum = 0
  for (i in 1:n) sum = sum + y[i]

  mean = sum / n

  return(mean)

}
```

## Use The Function

```{r}
x = rnorm(1000, mean = 23.42, sd = 4)

our_mean = my_mean_function(x)
our_mean

mean(x)
```
