---
title: "Introduction to R"
subtitle: "Advanced Statistics Lab"
author: "Claas Heuer"
format:
  revealjs:
    footer: "[Lab notes](../labs/r-intro.html)"
---

## Why R In This Lab?

R is the working language for reproducing the statistical examples.

We need it for:

- data inspection,
- model matrices,
- optimization,
- animal models,
- Bayesian models.

::: notes
Keep this pragmatic. The point is not to sell R as a language. The point is to remove friction before the statistical material begins.
:::

## The Course Pattern

1. Slides carry the narrative.
2. Lab notes carry runnable code.
3. RStudio is used for live execution.
4. Students rerun the same chunks locally.

## Setup Check

From the repository root:

```bash
Rscript setup/check.R --minimal
```

```{r}
library(ggplot2)
library(knitr)

theme_set(theme_minimal(base_size = 16))
```

## Objects

```{r}
milk_kg <- 31.4
fat_percent <- 4.2

fat_kg <- milk_kg * fat_percent / 100
fat_kg
```

Objects let us name intermediate results and reuse them.

## Vectors

```{r}
milk <- c(28.1, 31.4, 27.9, 35.2, 33.0)
herd <- c("A", "A", "B", "B", "B")

mean(milk)
table(herd)
milk[milk > 30]
```

## Data Frames

One row per record. One column per variable.

```{r}
animals <- data.frame(
  animal = paste0("cow_", 1:6),
  herd = c("A", "A", "B", "B", "C", "C"),
  sire = c("s1", "s1", "s2", "s2", "s3", "s3"),
  milk_kg = c(28.1, 31.4, 27.9, 35.2, 33.0, 30.5),
  fat_percent = c(4.1, 4.2, 3.8, 4.5, 4.0, 4.3)
)

animals
```

## Inspect First

```{r}
str(animals)
summary(animals)
```

Do this before modelling.

## Filter And Derive

```{r}
animals[animals$herd == "B", ]

animals$fat_kg <- animals$milk_kg * animals$fat_percent / 100
animals
```

## Files

```{r}
csv_file <- file.path(tempdir(), "animal_records.csv")

write.csv(animals, file = csv_file, row.names = FALSE)
animals_from_file <- read.csv(csv_file)

head(animals_from_file)
```

Always inspect after import.

## Matrices

```{r}
M <- matrix(
  c(
    0, 1, 2,
    1, 1, 0,
    2, 0, 1,
    0, 2, 1
  ),
  nrow = 4,
  byrow = TRUE
)

colnames(M) <- paste0("marker_", 1:3)
M
```

## Matrix Operations

```{r}
colMeans(M)
crossprod(M)
```

These operations reappear in regression, genomic prediction, and mixed models.

## Design Matrix

```{r}
X_herd <- model.matrix(~ herd, data = animals)
X_herd
```

A design matrix turns model structure into numbers.

## Visual Checks

```{r}
ggplot(animals, aes(milk_kg, fat_percent, color = herd)) +
  geom_point(size = 3) +
  labs(
    x = "Milk yield (kg/day)",
    y = "Fat (%)"
  )
```

## Functions

```{r}
my_mean <- function(y) {
  n <- length(y)
  sum_y <- sum(y)
  sum_y / n
}

my_mean(animals$milk_kg)
mean(animals$milk_kg)
```

## Why Functions Matter

Later in the course:

- an objective function is an R function,
- a likelihood is an R function,
- a simulation is usually a function,
- repeated model checks should be functions.

## Reproducible Script Pattern

1. Load packages.
2. Read or create data.
3. Inspect data.
4. Transform data.
5. Fit or compute.
6. Check and plot results.

## Take-Home

The first goal is not advanced R.

The goal is enough R to make the statistical workflow reproducible.
