---
title: "Introduction to R"
date: "`r Sys.Date()`"
author: Claas Heuer
output: rmarkdown::github_document
#output:
#  rmdformats::downcute:
#    self_contained: true
#    thumbnails: true
#    lightbox: true
#    gallery: false
#    highlight: tango
#    use_bookdown: false
    # Rscript -e 'library(rmarkdown); render("1_R_intro.rmd", output_file = "./out/1_R_intro.html")'
---


```{r, include = FALSE}
if(!require(pacman)) install.packages('pacman')
library(pacman)
p_load(data.table, ggplot2, rrBLUP, lattice, brms, knitr, plotly, prettydoc, lubridate, DT, lsmeans, car, htmltools, pander, tidyverse)
interactive = FALSE
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

```{r global_options, include=FALSE}

######################
### General Config ###
######################

knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
		      echo=FALSE, warning=FALSE, message=FALSE, include = FALSE, eval = TRUE)

fontSize = 15

```
# Get R

[From CRAN](https://cran.r-project.org/)

[Rstudio](https://www.rstudio.com/products/rstudio/download/#download)	

Ubuntu: `sudo apt install r-base`

# Some Introduction Material

From the developers: [Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf)	

Rstudio: [Documentation](https://docs.rstudio.com/)

[R for Data Science](https://r4ds.had.co.nz/introduction.html)

# Background

 - R is a general programming language with a specialization for statistics 
 - R is generally being used through a [REPL](https://en.wikipedia.org/wiki/Read%E2%80%93eval%E2%80%93print_loop) and interpreted live
 - Functional Programming paradigm: All about `objects` and `functions`
 - Major Appeals:
   - Extremely large ecosystem of packages, contributors, support system and companies
   - Many general and special solutions already available in R
   - Easy to learn and get going
   - Amongst the best graphics/plotting ecosystem for scientific output
 - Major Drawbacks: 
   - Functional Programming lends itself to "scripting" without structure
   - Barrier for introducing structure through packages high
   - R is an interpreted and not compiled language: no optimization 
   - "Always copy paradigm" makes R slow in many settings
   - R is old and its [successors](https://julialang.org/) already exist

# Basics

## Assigning and manipulating variables

Note: R has 2 assignment operator: `<-` and `=`. 
For all relevant porpuses of the tutorial, we can safely use either of those operators interchably.

```{r assign_varialbes, include = TRUE, echo = TRUE}

############################################
### Assigning and manipulating variables ###
############################################

x = 5
print(x)

# just givng the variable name, will print it to the REPL by default
x

# assign value from one variable to the next
y = x

# object did not trigger a copy of "x", rather y "points" to x
print(y)

# change the value of x
x = 3

print(x)

# at the moment we changed the value of "x", a copy of the old "x" was triggered and assigned to "y"
print(y)

```

## Working with data frames

One of the most important object types in R are `data.frames`. They are rectangular tables
with rows and columns, whereas the column can have different `types`, which is the main
differntiator to a `matrix`, where the column types are all identical.

```{r data.frames, include = TRUE, echo = TRUE}

############################################
### Assigning and manipulating variables ###
############################################

dat = data.frame(
		 name = c("Adam", "Paul", "Elly"),
		 age = c(22, 43, 54)
		 )

# look at general structure
str(x)

# extracting single column and storing in a vector object
names = dat$name

# can also be accessed with the bracket operator
ages = dat[, "age"]

str(names)
str(ages)

# subsetting
dat[dat$name == "Paul",]
dat[dat$age %in% c(22, 54),]

# adding a column to a data frame
dat$id = 1:nrow(dat)
dat[, "id_2"] = dat$id + 2

print(dat)

# conditional assignment
dat$id[dat$name == "Adam"] <- 12

print(dat)
```
## Reading and Writing Data from/to files

Most of the time, data is being innput to R from files.
However, R has interfaces to most data storage systems out there,
including all databases and cloud storages.


```{r files, include = TRUE, echo = TRUE}

###########################
### Read and Write data ###
###########################

# create data frame
dat = data.frame(
		 name = c("Adam", "Paul", "Elly"),
		 age = c(22, 43, 54)
		 )
str(dat)

# write data to disk as csv
write.table(dat, file = "my_data.csv", row.names = FALSE, sep = ",")

# read file back in
dat_read = read.table(file = "my_data.csv", sep = ",", header = TRUE)
str(dat_read)

```

## Vectors and Matrices

In R, everything is a `vector` or a `list` or an `array` of vectors.
A `data.frame` is a list of vectors of different types, a `matrix` is a list
of vectors of a single type. A `scalar` is a vector of length 1.

```{r vectors, include = TRUE, echo = TRUE}

###########################
### Vectos and Matrices ###
###########################

# integer vector of length one
x = as.integer(1)
str(x)
is.vector(x)

# numeric vector
y = rnorm(10)
str(x)
x

#  matrix
D = matrix(data = rnorm(5*5), nrow = 5, ncol = 5)
D
class(D)

# specialized functions can be applied to matrices
means = colMeans(D)
means

# extract diagonal
d = diag(D)
d

```

## Data Exploration and Visualization

Getting a good overview and understanding of the data sets we work with is very important.
Summary statistics and visualizations are important tools here.
We will be using [dlookr](https://github.com/choonghyunryu/dlookr) and [ggplot2](https://ggplot2.tidyverse.org/).

```{r visuals, include = TRUE, echo = TRUE}

##########################################
### Data Exploration and Visualization ###
##########################################

if(!require(dlookr)) install.packages('dlookr')
if(!require(ggplot2)) install.packages('ggplot2')
library(dlookr, ggplot2)

# load a public data set
data(iris)
str(iris)

# broad overview of data set
dlookr::diagnose(iris)

# plot our data
# scatterplot
p = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
	geom_point()

p

# histogram for temp by month
p2 = ggplot(iris, aes(x = Petal.Length)) + 
	geom_histogram(aes(fill = Species), bins = 20) + 
	facet_wrap(~ Species) 

p2

```

## Writing Functions

All operators in R are functions, for instance `+` is just a special way
of calling a function:

```{r function, include = TRUE, echo = TRUE, eval = FALSE}

# get help for a function
?sum

# get help for the function '+'
?`+`

# check the function definition for `+`
fix(`+`)
```

A function is an operation on input variables ($f(x)), the function can be anything
interpretable by R. For example, this is how we can define a function that computes
and returns the empirical mean (average) of a numeric vector:

```{r function_mean, include = TRUE, echo = TRUE}

my_mean_function = function(y) {

	n = length(y)
	sum = 0
	for(i in 1:n) sum = sum + y[i]

	mean = sum / n

	return(mean)

}
```

We can now use that function inside R for estimating means on any input vector `y` for which
the the operators `+` and `/` are defined.

```{r mean_call, include = TRUE, echo = TRUE}

# generate random vector
x = rnorm(1000, mean = 23.42, sd = 4)

# calculate mean with our function
our_mean = my_mean_function(x)
our_mean

# compare with R's built in function
mean(x)
```
