---
title: "Optimization"
date: "`r Sys.Date()`"
author: Claas Heuer
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: false
    # Rscript -e 'library(rmarkdown); render("2_Optimization.rmd", output_file = "./out/2_Optimization.html")'
---

```{r, include = FALSE}
if(!require(pacman)) install.packages('pacman')
library(pacman)
p_load(data.table, ggplot2, rrBLUP, lattice, brms, knitr, plotly, prettydoc, lubridate, DT, lsmeans, car, htmltools, pander, tidyverse)
interactive = FALSE
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

```{r global_options, include=FALSE}

######################
### General Config ###
######################

knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
		      echo=FALSE, warning=FALSE, message=FALSE, include = FALSE, eval = TRUE)

fontSize = 15

```

  
# Optimization - Introduction

Optimization deals with finding values for variables that maximize or minimze an
objective function of those variables.

The general formulation is:

$$
Objective Function = max(f(x)),
$$

where $x$ is a vecvtor of variables and $f(x)$ is a function that takes the vector of varialbes $x$
as input.

## Example

Our optimization function might look like this:

$$
f(x) = (x_{1}^2 + x_2) - (x_{2}^{2})
$$

Our vector of variables contains 2 elements, $x_1$ and $x_2$ and we want to maximize
the sum of the first variable squared plut the second variable minus the square of the second variable.

Most of the time, our variables are constrained in the sense that they can only take values
in a given range. Constraints essentially narrow down the possible solution space. For real
valued variables, the precision matters as well.
For our optimization function, we might want to impose the constraint for the variables
to lye between 1 and 100:

$$
\begin{aligned}                                                                                          
x_1 &>= 1 \\
x_2 &>= 1 \\
x_1 &<= 100 \\
x_2 &<= 100
\end{aligned}                                                                                          
$$

This defines our solution space and, assuming there is no closed form analytical solution, we can find the optimal solution
by evaluating the objective function for all possible realisations of $x_1$ and $x_2$.

Example in R

```r
library(lattice)
lower = 1
upper = 100
x1 = lower:upper
x2 = lower:upper

# loop over all possible combinations and evaluate our function
sol = outer(
	    X = x1, 
	    Y = x2,
	    FUN = function(x,y) (x**2 + y) - y**2
	   )

# grid search solutions
levelplot(sol)

# optimum solution(s)
which(sol == max(sol), arr.ind = TRUE)
```

Finding the optimum solution through a grid search reaches its limits very fast and we
have to use algorithms that find the global maximum (or minimum) without having to evaluate
the objective function across the entire solution space.

In R the most generic way of using algorithms for optimzation problems is `optim`.
For our problem, it would look like this

```r
solOptim = optim(
		 par = c(1,1),
		 fn = function(par) (par[1]**2 + par[2]) - par[2]**2,
		 lower = 1,
		 upper = 100,
		 method = "L-BFGS-B",
		 control = list(fnscale = -1) # maximize
		)

# check solution
solOptim$par
```

The optimization yields the same solution as our grid search.
One important thing to note here is how the constraints enter 
the optimzation algorithm. It might be necessary to incorporate 
the constraints into the objective function, if the optimization algorithm
does not allow for constraint parameter space. In our example,
this could look like this:

```r
objective = function(par, lower, upper) {

# constraints
	if(
	   par[1] < lower |
	   par[2] < lower |
	   par[1] > upper |
	   par[2] > upper 
	  ) {

		return(0)

	} else {

		return((par[1]**2 + par[2]) - par[2]**2)

	}

}

solOptim = optim(
		 par = c(1,1),
		 fn = function(par) objective(par, lower = 1, upper = 100),
		 method = "L-BFGS", # this algorithm does not allow for constraint variables
		 control = list(fnscale = -1) # maximize
		)

# check solution - very close to the strict integer solution
solOptim$par
```


