---
title: "Bayesian Inference"
subtitle: "Priors, brms, and animal-model patterns"
format:
  html:
    code-fold: false
---

<div class="view-switch"><a href="../lectures/bayesian-inference.html">Open slides</a></div>

Bayesian inference combines the likelihood with prior assumptions. In this lab the focus is practical: specify priors, fit models with `brms`, update priors from previous evidence, and recognize the animal-model structure in a Bayesian formulation.

```{r}
#| include: false
library(ggplot2)
library(knitr)

theme_set(theme_minimal(base_size = 13))
```

## Bayes' Theorem

$$
p(\theta \mid y)
=
\frac{p(y \mid \theta)p(\theta)}{p(y)}.
$$

For model fitting, the denominator is a scaling constant:

$$
p(\theta \mid y)
\propto
p(y \mid \theta)p(\theta).
$$

That is enough for this lab: likelihood times prior gives the fitted Bayesian model.

## Estimating A Normal Mean

The original first example estimates the mean of `Sepal.Width` in the `iris` data:

$$
\mathbf{y} \sim N(\mu, \sigma_e^2),
\qquad
\mu \sim N(\mu_\text{prior}, \sigma_\text{prior}^2).
$$

The key teaching point is the role of the prior location and prior standard deviation.

```{r}
data(iris)

iris_mean <- data.frame(
  trait = "Sepal.Width",
  n = length(iris$Sepal.Width),
  mean = mean(iris$Sepal.Width),
  sd = sd(iris$Sepal.Width)
)

kable(iris_mean, digits = 3)
```

## brms Mean Example

This is the in-class code pattern from the old lab. It is not evaluated during site rendering, because Stan compilation should not block the course website build.

```{r eval=FALSE}
library(brms)
library(posterior)

data(iris)

# inspect default priors
get_prior(Sepal.Width ~ 1, data = iris)

# custom prior for the intercept
priors <- set_prior("normal(0, 0.2)", class = "Intercept")

fit <- brm(
  Sepal.Width ~ 1,
  data = iris,
  family = gaussian(),
  prior = priors,
  sample_prior = TRUE,
  backend = "cmdstanr",
  chains = 2,
  iter = 1000,
  seed = 2026,
  refresh = 0
)

draws <- as_draws_array(fit)[, 1, , drop = TRUE]

samples <- rbind(
  data.frame(type = "prior", samples = as.numeric(draws[, "prior_Intercept"])),
  data.frame(type = "posterior", samples = as.numeric(draws[, "b_Intercept"]))
)

aggregate(samples ~ type, data = samples, FUN = function(x) c(mean = mean(x), sd = sd(x)))
plot(fit)
```

For the rendered notes, we use a lightweight preview of the same prior-vs-estimate idea.

```{r}
set.seed(2026)

prior_samples <- rnorm(4000, mean = 0, sd = 0.2)
estimate_samples <- rnorm(
  4000,
  mean = mean(iris$Sepal.Width),
  sd = sd(iris$Sepal.Width) / sqrt(nrow(iris))
)

samples <- rbind(
  data.frame(type = "prior", samples = prior_samples),
  data.frame(type = "estimate from data", samples = estimate_samples)
)

sample_summary <- do.call(
  rbind,
  lapply(split(samples$samples, samples$type), function(x) {
    data.frame(n = length(x), mean = mean(x), sd = sd(x))
  })
)
sample_summary$type <- rownames(sample_summary)
rownames(sample_summary) <- NULL

kable(
  sample_summary[, c("type", "n", "mean", "sd")],
  digits = 3
)
```

```{r}
ggplot(samples, aes(samples, fill = type)) +
  geom_density(alpha = 0.55) +
  scale_fill_manual(values = c("prior" = "#526A83", "estimate from data" = "#68855C")) +
  labs(x = "Mean of Sepal.Width", y = "Density", fill = NULL)
```

## Updating Priors From Existing Evidence

The old lab used a first experiment to inform the prior for a second experiment. That is the practical Bayesian workflow worth keeping.

```{r eval=FALSE}
library(brms)
library(posterior)

data(iris)

iris_first <- iris[1:50, ]
iris_second <- iris[51:nrow(iris), ]

fit_first <- brm(
  Sepal.Width ~ 1,
  data = iris_first,
  family = gaussian(),
  prior = set_prior("normal(0, 4)", class = "Intercept"),
  sample_prior = TRUE,
  backend = "cmdstanr",
  chains = 2,
  iter = 1000,
  seed = 2026,
  refresh = 0
)

draws_first <- as_draws_array(fit_first)[, 1, , drop = TRUE]
mu_prior <- mean(draws_first[, "b_Intercept"])
sd_prior <- sd(draws_first[, "b_Intercept"])

fit_second <- brm(
  Sepal.Width ~ 1,
  data = iris_second,
  family = gaussian(),
  prior = set_prior(paste0("normal(", mu_prior, ",", sd_prior, ")"), class = "Intercept"),
  sample_prior = TRUE,
  backend = "cmdstanr",
  chains = 2,
  iter = 1000,
  seed = 2026,
  refresh = 0
)
```

The same workflow can be demonstrated quickly with normal approximations:

```{r}
update_mean_preview <- function(y, prior_mean, prior_sd) {
  data_sd <- sd(y)
  n <- length(y)
  precision <- 1 / prior_sd^2 + n / data_sd^2
  post_sd <- sqrt(1 / precision)
  post_mean <- (prior_mean / prior_sd^2 + n * mean(y) / data_sd^2) / precision
  c(mean = post_mean, sd = post_sd)
}

iris_first <- iris[1:50, ]
iris_second <- iris[51:nrow(iris), ]

first <- update_mean_preview(iris_first$Sepal.Width, prior_mean = 0, prior_sd = 4)
second <- update_mean_preview(iris_second$Sepal.Width, prior_mean = first["mean"], prior_sd = first["sd"])

kable(
  data.frame(
    stage = c("wide prior", "after first subset", "after second subset"),
    mean = c(0, unname(first["mean"]), unname(second["mean"])),
    sd = c(4, unname(first["sd"]), unname(second["sd"]))
  ),
  digits = 3
)
```

## Multivariate Animal Model In brms

The old chapter ended with the structure of a multivariate animal model. The important model form is:

$$
\mathbf{Y}
=
\mathbf{X}\mathbf{B}
+ \mathbf{Z}\mathbf{U}
+ \mathbf{E}.
$$

For two traits, `brms` can express this through `mvbind()` and covariance-aware grouping terms:

```{r eval=FALSE}
library(brms)
library(cmdstanr)
library(BGLR)
library(Matrix)

data(milk, package = "BGLR")

# P, A, D are prepared from the pedigree and milk records.
# A is the additive relationship matrix for the animals in D.

brmf <- brmsformula(
  mvbind(milk, fat) ~ 1 +
    factor(lact) +
    herd +
    (1 | g | gr(idA, cov = A)) +
    (1 | p | gr(id)),
  data = D,
  data2 = list(A = A)
)

get_prior(brmf, data = D, data2 = list(A = A), family = gaussian())

stan_code <- make_stancode(
  formula = brmf,
  data = D,
  data2 = list(A = A),
  family = gaussian()
)

mod <- brm(
  formula = brmf,
  data = D,
  data2 = list(A = A),
  family = gaussian(),
  backend = "cmdstanr",
  chains = 4,
  cores = 4,
  iter = 2000
)

vcp <- VarCorr(mod, summary = FALSE)
G <- vcp$idA$cov
PE <- vcp$id$cov
R <- vcp$residual$cov
P <- G + PE + R
```

The point here is not to run a large animal model live in the first Bayesian session. The point is to see how the same model components from the animal-model chapter map into a Bayesian `brms` formula.

## Cached Artifacts

For class, expensive Stan fits should be generated ahead of time:

```bash
Rscript scripts/build-model-artifacts.R --stan
```

Then the lab can load `.rds` artifacts from `artifacts/models/` instead of fitting live.

```{r}
artifact_file <- "artifacts/models/bayesian_mean_brms.rds"

if (file.exists(artifact_file)) {
  artifact <- readRDS(artifact_file)
  artifact$summary
} else {
  "No cached brms artifact found. Use scripts/build-model-artifacts.R --stan to generate it."
}
```

## Resources

- [Stan](https://mc-stan.org/)
- [brms](https://github.com/paul-buerkner/brms)
- [AnimalModels](https://github.com/cheuerde/AnimalModels)
- [BayesianAlphabet](https://github.com/cheuerde/BayesianAlphabet)

## Exercises

1. Change the prior standard deviation in the `brms` mean example.
2. Run the first `brms` model after checking the Stan setup.
3. Use the first 75 iris rows as the first experiment and the remaining rows as the second.
4. Inspect `make_stancode()` for the multivariate animal-model formula.
5. Write down which variance-covariance matrices are represented by `G`, `PE`, and `R`.
