---
title: "Least Squares"
subtitle: "Visual fitting, closed-form solutions, and shrinkage"
format:
  html:
    code-fold: false
---

<div class="view-switch"><a href="../lectures/linear-regression.html">Open slides</a></div>

Least squares is the first statistical optimization problem in the course. We choose a line by making the vertical distances from the observed points to the line as small as possible.

```{r}
#| include: false
library(ggplot2)
library(knitr)

theme_set(theme_minimal(base_size = 13))
```

## A Line Through Data

Assume we record concentrate intake and milk yield for a group of cows. We want a simple model for expected milk yield:

$$
\widehat{y}_i = \beta_0 + \beta_1 x_i.
$$

```{r}
set.seed(2026)

milk_data <- data.frame(
  feed_kg = seq(4, 12, length.out = 36)
)

milk_data$milk_kg <- 17.5 + 1.65 * milk_data$feed_kg +
  rnorm(nrow(milk_data), sd = 1.8)

kable(head(milk_data), digits = 2)
```

Start with a candidate line. The vertical distances from each point to that line are the residuals.

```{r}
candidate_beta <- c(intercept = 18, slope = 1.5)

milk_data$candidate_pred <- candidate_beta["intercept"] +
  candidate_beta["slope"] * milk_data$feed_kg

ggplot(milk_data, aes(feed_kg, milk_kg)) +
  geom_segment(
    aes(xend = feed_kg, yend = candidate_pred),
    color = "#b85c38",
    linewidth = 0.45,
    alpha = 0.75
  ) +
  geom_point(size = 2.4, color = "#155f7a") +
  geom_line(aes(y = candidate_pred), color = "#8a4f12", linewidth = 1) +
  labs(
    x = "Concentrate intake (kg/day)",
    y = "Milk yield (kg/day)"
  )
```

The least-squares line is the line for which the sum of squared vertical distances is smallest.

## The Least-Squares Problem

For a simple line, the unknown parameter vector is

$$
\boldsymbol{\beta} =
\begin{bmatrix}
\beta_0 \\
\beta_1
\end{bmatrix},
$$

where $\beta_0$ is the intercept and $\beta_1$ is the slope. The objective is the sum of squared errors:

$$
\operatorname{SSE}(\boldsymbol{\beta})
=
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \beta_1 x_i\right)^2.
$$

The optimization problem is

$$
\widehat{\boldsymbol{\beta}}
=
\operatorname*{arg\,min}_{\boldsymbol{\beta} \in \mathbb{R}^2}
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \beta_1 x_i\right)^2.
$$

```{r}
sse_feed_model <- function(par, data) {
  beta_0 <- par[1]
  beta_1 <- par[2]

  y_hat <- beta_0 + beta_1 * data$feed_kg
  sum((data$milk_kg - y_hat)^2)
}

sse_feed_model(candidate_beta, milk_data)
```

## Grid Search

We can use a grid search to see the objective function. Each point in this grid is one possible line.

```{r}
loss_grid <- expand.grid(
  intercept = seq(8, 28, length.out = 90),
  slope = seq(0.4, 2.8, length.out = 90)
)

loss_grid$sse <- mapply(
  FUN = function(intercept, slope) sse_feed_model(c(intercept, slope), milk_data),
  intercept = loss_grid$intercept,
  slope = loss_grid$slope
)

best_grid <- loss_grid[which.min(loss_grid$sse), ]

kable(best_grid, digits = 3)
```

```{r}
ggplot(loss_grid, aes(intercept, slope, z = sse)) +
  geom_contour_filled(bins = 18) +
  geom_point(
    data = best_grid,
    aes(intercept, slope),
    inherit.aes = FALSE,
    color = "white",
    size = 3
  ) +
  labs(
    x = "Intercept",
    y = "Feed response",
    fill = "SSE"
  )
```

## Solving It With `optim()`

Grid search is useful for teaching, but it becomes inefficient quickly. `optim()` searches the parameter space directly.

```{r}
fit_optim <- optim(
  par = c(15, 1),
  fn = sse_feed_model,
  data = milk_data,
  method = "BFGS"
)

coef_optim <- setNames(fit_optim$par, c("intercept", "slope"))

kable(
  data.frame(
    parameter = names(coef_optim),
    estimate = as.numeric(coef_optim)
  ),
  digits = 4
)
```

```{r}
milk_data$optim_pred <- coef_optim["intercept"] + coef_optim["slope"] * milk_data$feed_kg

ggplot(milk_data, aes(feed_kg, milk_kg)) +
  geom_segment(
    aes(xend = feed_kg, yend = optim_pred),
    color = "#b85c38",
    linewidth = 0.45,
    alpha = 0.55
  ) +
  geom_point(size = 2.4, color = "#155f7a") +
  geom_line(aes(y = optim_pred), color = "#8a4f12", linewidth = 1) +
  labs(
    x = "Concentrate intake (kg/day)",
    y = "Milk yield (kg/day)"
  )
```

## Bound Constraints

Some parameters have natural bounds. For example, a response to additional feed may be constrained to be non-negative in a simple teaching model:

$$
\begin{aligned}
\widehat{\boldsymbol{\beta}}
&=
\operatorname*{arg\,min}_{\boldsymbol{\beta} \in \mathbb{R}^2}
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \beta_1 x_i\right)^2 \\
\text{subject to } 0 &\le \beta_1 \le 4.
\end{aligned}
$$

```{r}
fit_bounded <- optim(
  par = c(15, 1),
  fn = sse_feed_model,
  data = milk_data,
  method = "L-BFGS-B",
  lower = c(-Inf, 0),
  upper = c(Inf, 4)
)

kable(
  data.frame(
    parameter = c("intercept", "slope"),
    unconstrained = coef_optim,
    bounded = fit_bounded$par
  ),
  digits = 4
)
```

Constraints are model assumptions. They can stabilize estimation, but they can also hide problems if the constraint is biologically wrong or too strict.

```{r}
fit_too_tight <- optim(
  par = c(15, 1),
  fn = sse_feed_model,
  data = milk_data,
  method = "L-BFGS-B",
  lower = c(-Inf, 0),
  upper = c(Inf, 1)
)

kable(
  data.frame(
    parameter = c("intercept", "slope"),
    unconstrained = coef_optim,
    too_tight = fit_too_tight$par
  ),
  digits = 4
)
```

When an estimate lands exactly on a boundary, the boundary itself becomes part of the result and needs interpretation.

## Matrix Form

The same model can be written for all observations at once:

$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{e}.
$$

For the simple line, $\mathbf{X}$ has one column of ones for the intercept and one column containing feed intake:

```{r}
y_feed <- milk_data$milk_kg
X_feed <- model.matrix(~ feed_kg, data = milk_data)

kable(head(X_feed), digits = 2)
```

The least-squares objective is

$$
\widehat{\boldsymbol{\beta}}
=
\operatorname*{arg\,min}_{\boldsymbol{\beta}}
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})'
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta}).
$$

## Closed-Form Solution

For ordinary least squares, we do not need a general optimizer. The normal equations are

$$
\mathbf{X}'\mathbf{X}\widehat{\boldsymbol{\beta}}
=
\mathbf{X}'\mathbf{y}.
$$

If $\mathbf{X}'\mathbf{X}$ can be inverted, the closed-form solution is

$$
\widehat{\boldsymbol{\beta}}
=
(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}.
$$

```{r}
beta_closed <- solve(t(X_feed) %*% X_feed, t(X_feed) %*% y_feed)
coef_lm <- coef(lm(milk_kg ~ feed_kg, data = milk_data))

comparison <- data.frame(
  method = c("optim", "closed form", "lm"),
  intercept = c(coef_optim["intercept"], beta_closed[1, 1], coef_lm[[1]]),
  slope = c(coef_optim["slope"], beta_closed[2, 1], coef_lm[["feed_kg"]])
)

kable(comparison, digits = 4)
```

`optim()`, the closed-form solution, and `lm()` agree because they solve the same least-squares problem.

## Intercept-Only Model

The intercept-only model is the smallest possible least-squares model:

$$
y_i = \beta_0 + e_i.
$$

```{r}
X_mean <- matrix(1, nrow = length(y_feed), ncol = 1)
colnames(X_mean) <- "Intercept"

beta_mean <- solve(t(X_mean) %*% X_mean, t(X_mean) %*% y_feed)

kable(
  data.frame(
    estimate = as.numeric(beta_mean),
    empirical_mean = mean(y_feed)
  ),
  digits = 3
)
```

The intercept-only least-squares estimate is the sample mean.

## More Design Matrices

Design matrices can contain continuous predictors, factors, interactions, and transformations. Here is the same least-squares machinery on the built-in `airquality` data:

```{r}
data(airquality)

dat <- na.omit(airquality[, c("Wind", "Temp", "Month")])
y <- dat$Wind
X_temp <- model.matrix(~ Temp, data = dat)

beta_temp <- solve(t(X_temp) %*% X_temp, t(X_temp) %*% y)

kable(
  data.frame(
    parameter = rownames(beta_temp),
    estimate = as.numeric(beta_temp)
  ),
  digits = 3
)
```

```{r}
dat$pred_lm <- drop(X_temp %*% beta_temp)

ggplot(dat, aes(Temp, Wind)) +
  geom_point(color = "#0c6773", size = 2) +
  geom_line(aes(y = pred_lm), color = "#c46a2b", linewidth = 1) +
  labs(x = "Temperature", y = "Wind")
```

## Multiple Regression

```{r}
X_multi <- model.matrix(~ Temp + Month, data = dat)
beta_multi <- solve(t(X_multi) %*% X_multi, t(X_multi) %*% y)

kable(
  data.frame(parameter = rownames(beta_multi), estimate = as.numeric(beta_multi)),
  digits = 3
)
```

## Ridge Regression

Ridge regression adds a penalty to shrink coefficients:

$$
\widehat{\boldsymbol{\beta}}_{\lambda}
=
\operatorname*{arg\,min}_{\boldsymbol{\beta}}
\left[
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})'
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})
+ \lambda \boldsymbol{\beta}'\boldsymbol{\beta}
\right].
$$

The closed-form ridge solution is

$$
\widehat{\boldsymbol{\beta}}_{\lambda}
=
(\mathbf{X}'\mathbf{X} + \lambda\mathbf{I})^{-1}\mathbf{X}'\mathbf{y}.
$$

```{r}
set.seed(2026)
n <- 80
p <- 20
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
beta_true <- c(2.5, -2, 1.2, rep(0, p - 3))
y_ridge <- 12 + X %*% beta_true + rnorm(n, sd = 3)

X_scaled <- scale(X)
X_ridge <- cbind(Intercept = 1, X_scaled)

ridge_fit <- function(lambda, X, y) {
  penalty <- diag(ncol(X))
  penalty[1, 1] <- 0
  solve(t(X) %*% X + lambda * penalty, t(X) %*% y)
}

lambdas <- c(0, 1, 10, 100)
ridge_out <- do.call(
  rbind,
  lapply(lambdas, function(lambda) {
    beta <- ridge_fit(lambda, X_ridge, y_ridge)
    data.frame(lambda = lambda, coefficient = seq_along(beta) - 1, estimate = as.numeric(beta))
  })
)

ggplot(subset(ridge_out, coefficient > 0), aes(lambda, estimate, group = coefficient)) +
  geom_line(alpha = 0.45, color = "#0c6773") +
  scale_x_log10(breaks = lambdas[-1]) +
  labs(x = expression(lambda), y = "Coefficient estimate")
```

## Bridge To Animal Models

Ridge regression is the fixed-design version of shrinkage. Animal models shrink random effects through covariance assumptions:

$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{Z}\mathbf{u} + \mathbf{e},
\qquad
\mathbf{u} \sim N(\mathbf{0}, \mathbf{G}),
\qquad
\mathbf{e} \sim N(\mathbf{0}, \mathbf{R}).
$$

If the variance components are known, the BLUP solution can be written as a penalized least-squares problem:

$$
\begin{aligned}
\widehat{\boldsymbol{\beta}}, \widehat{\mathbf{u}}
&=
\operatorname*{arg\,min}_{\boldsymbol{\beta}, \mathbf{u}}
\Big[
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta} - \mathbf{Z}\mathbf{u})'
\mathbf{R}^{-1}
(\mathbf{y} - \mathbf{X}\boldsymbol{\beta} - \mathbf{Z}\mathbf{u}) \\
&\qquad\qquad\qquad\qquad
+ \mathbf{u}'\mathbf{G}^{-1}\mathbf{u}
\Big].
\end{aligned}
$$

For a small teaching version with unrelated animals and known $\lambda = \sigma_e^2 / \sigma_a^2$, this reduces to the mixed model equations:

$$
\begin{bmatrix}
\mathbf{X}'\mathbf{X} & \mathbf{X}'\mathbf{Z} \\
\mathbf{Z}'\mathbf{X} & \mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}
\end{bmatrix}
\begin{bmatrix}
\widehat{\boldsymbol{\beta}} \\
\widehat{\mathbf{u}}
\end{bmatrix}
=
\begin{bmatrix}
\mathbf{X}'\mathbf{y} \\
\mathbf{Z}'\mathbf{y}
\end{bmatrix}.
$$

```{r}
set.seed(2026)

n_animals <- 12
records_per_animal <- 2
animal_id <- factor(rep(seq_len(n_animals), each = records_per_animal))

true_u <- rnorm(n_animals, mean = 0, sd = 2.0)
animal_y <- 30 + true_u[animal_id] + rnorm(length(animal_id), sd = 3.0)

X_animal <- matrix(1, nrow = length(animal_y), ncol = 1)
Z_animal <- model.matrix(~ animal_id - 1)
lambda <- 3.0^2 / 2.0^2

lhs <- rbind(
  cbind(t(X_animal) %*% X_animal, t(X_animal) %*% Z_animal),
  cbind(t(Z_animal) %*% X_animal, t(Z_animal) %*% Z_animal + lambda * diag(n_animals))
)

rhs <- rbind(t(X_animal) %*% animal_y, t(Z_animal) %*% animal_y)
solution <- solve(lhs, rhs)

animal_model_example <- data.frame(
  animal = seq_len(n_animals),
  true_effect = true_u,
  blup = solution[-1, 1]
)

kable(head(animal_model_example), digits = 3)
```

## Exercises

1. Change the candidate intercept and slope in the first figure. How do the residuals change?
2. Make the upper bound on the slope smaller than the unconstrained estimate. How can you tell that the solution is boundary-driven?
3. Compare `solve(t(X) %*% X, t(X) %*% y)` with `lm()`.
4. Add `Solar.R` to the multiple regression design matrix.
5. Standardize `Temp` manually and refit the simple regression.
