---
title: "Maximum Likelihood"
subtitle: "Estimating parameters by optimizing probability"
format:
  html:
    code-fold: false
---

<div class="view-switch"><a href="../lectures/maximum-likelihood.html">Open slides</a></div>

Maximum likelihood estimates the parameter values that make the observed data most probable under a statistical model. The implementation pattern in this chapter is deliberately explicit: construct the likelihood, collect the parameters in one vector, and pass that vector to `optim()`.

## Definition

For independent observations $y_1,\dots,y_n$ with density $p(y_i \mid \theta)$, the likelihood is

$$
L(\theta; \mathbf{y}) = \prod_{i=1}^{n} p(y_i \mid \theta).
$$

The maximum likelihood estimate is

$$
\widehat{\theta}
=
\operatorname*{arg\,max}_{\theta \in \Theta}
L(\theta; \mathbf{y}).
$$

The product can become extremely small. Therefore we usually optimize the log-likelihood:

$$
\log L(\theta; \mathbf{y})
=
\sum_{i=1}^{n} \log p(y_i \mid \theta).
$$

Because `optim()` minimizes, the function we pass to R is usually the negative log-likelihood:

$$
\widehat{\theta}
=
\operatorname*{arg\,min}_{\theta \in \Theta}
\left[-\sum_{i=1}^{n}\log p(y_i \mid \theta)\right].
$$

```{r}
#| include: false
library(ggplot2)
library(knitr)

theme_set(theme_minimal(base_size = 13))
```

## Normal Distribution

The normal density is

$$
f(y) = \frac{1}{\sigma \sqrt{2\pi}}
\exp\left[-\frac{1}{2}\left(\frac{y - \mu}{\sigma}\right)^2\right].
$$

The normal distribution has two parameters: the mean $\mu$ and the standard deviation $\sigma$.

```{r}
curve(
  dnorm(
    x,
    mean = 10,
    sd = 2
  ),
  from = 0,
  to = 20,
  main = "Normal distribution"
)
```

## Finding The Mean

Here $\sigma$ is fixed at 1 and only $\mu$ is estimated.

```{r}
set.seed(2026)

# generate random normal variable
mean = 4.5
sd = 1
n = 1000

y = rnorm(n, mean, sd)

# the likelihood function = L(y;.) = y ~ N(mu, sigma)
lik <- function(par, y, sd = 1) {

  # the parameter we want to estimate
  mean = par[1]

  # the likelihood of y
  L <- dnorm(x = y, mean = mean, sd = sd, log = TRUE)

  # the joint likelihood
  return(-sum(L))

}


# the parameters we wish to estimate
p <- c(0)
names(p) <- c("mu")

# Get the Maximum Likelihood estimates for our unknowns
ML <- optim(p, lik, y = y)

# compare ML results to empirical mean
print("Maximum Likelihood:")
ML$par

print("Empirical:")
mean(y)
```

## Finding Mean And Standard Deviation

Now estimate $\mu$ and $\sigma$ together. The only extra point is that $\sigma$ must be positive, so the likelihood returns `Inf` for invalid values.

```{r}
# likelihood for a positive standard deviation
sdlik <- function(x) {

  res <- 0
  if (x > 0) res = 1

  return(log(res))

}

# the likelihood function = L(y;.) = y ~ N(mu, sigma) * sigma ~ U(0, Inf)
lik <- function(par, y) {

  # the parameters we want to estimate
  mean = par[1]
  sd = par[2]

  # the likelihood of the standard deviation
  L2 <- sdlik(sd)

  # if sd is not allowed, stop here
  if (!is.finite(L2)) return(Inf)

  # the likelihood of y
  L1 <- dnorm(x = y, mean = mean, sd = sd, log = TRUE)

  # the joint likelihood
  return(-sum(L1 + L2))

}


# the parameters we wish to estimate
p <- c(0, 1)
names(p) <- c("mu", "sigma")

# Get the Maximum Likelihood estimates for our unknowns
ML <- optim(p, lik, y = y, method = "Nelder-Mead")

# compare ML results to empiricals
out <- rbind(
  ML = ML$par,
  empirical = c(mean(y), sd(y))
)

out
```

## Estimating Correlation

Now the likelihood is multivariate normal. The parameters are two means, two variances, and the correlation.

```{r}
library(mvtnorm)

# likelihood on correlation coefficient.
# this is essentially a uniform prior on
# the range -1 to 1 and 0 density outside
# that range
corlik <- function(x) {

  res <- 0
  if (x >= -1 & x <= 1) res = 1

  return(log(res))

}

# the likelihood for the variance components.
# uniform above 0 and zero if negative
varlik <- function(x) {

  res <- 0
  if (x > 0) res = 1

  return(log(res))

}

# the likelihood function = L(y;.) = y ~ MVN(mu, sigma) * rho ~ U(-1,1)
lik <- function(par, y) {

  # the prior = likelihood of rho (correlation coefficient)
  L2 <- corlik(par[5])

  # and the priors on the variance components
  L3 <- varlik(par[3])
  L4 <- varlik(par[4])

  # if any parameter is outside the allowed range, stop here
  if (!all(is.finite(c(L2, L3, L4)))) return(Inf)

  # we construct the covariance matrix out of the correlation coefficient
  # and the variance components
  covAB <- par[5] * (sqrt(par[3]) * sqrt(par[4]))
  sigma <- array(c(par[3], covAB, covAB, par[4]), dim = c(2, 2))

  # the likelihood of y
  L1 <- tryCatch(
    dmvnorm(x = y, mean = c(par[1], par[2]), sigma = sigma, log = TRUE),
    error = function(e) -Inf
  )

  if (!all(is.finite(L1))) return(Inf)

  # the joint likelihood
  return(-sum(L1 + L2 + L3 + L4))

}


# those are the optimizers available
methods = c("Nelder-Mead", "BFGS", "CG", "L-BFGS-B", "SANN", "Brent")

# the parameters we wish to estimate
p <- c(0, 0, 1, 1, 0.2)
names(p) <- c("mu1", "mu2", "tauA", "tauB", "rho")

# create random data
set.seed(2026)
n = 10
tauA <- 3
tauB <- 10
rho = 0.7
covAB <- rho * (sqrt(tauA) * sqrt(tauB))
muA <- 5
muB <- 20

# the variance-covariance matrix
sigma <- array(c(tauA, covAB, covAB, tauB), dim = c(2, 2))

# the random data
Y <- rmvnorm(n, c(muA, muB), sigma)

# Get the Maximum Likelihood estimates for our unknowns
ML <- optim(p, lik, y = Y, method = methods[1])

res <- as.list(ML$par)
res$covAB <- res$rho * (sqrt(res$tauA) * sqrt(res$tauB))

# compare ML estimates with empiricals
out <- rbind(
  unlist(res),
  c(colMeans(Y), var(Y[, 1]), var(Y[, 2]), cor(Y)[1, 2], cov(Y)[1, 2])
)
colnames(out) <- names(res)
rownames(out) <- c("ML", "empirical")

# inspect results
out
```

## Animal Model Likelihood

The animal model is

$$
\mathbf{y} = \mathbf{X}\mathbf{b} + \mathbf{a} + \mathbf{e},
$$

with the marginal distribution

$$
\mathbf{y} \sim MVN(\mathbf{X}\mathbf{b}, \mathbf{A}\sigma_a^2 + \mathbf{I}\sigma_e^2).
$$

The implementation below keeps the construction explicit. We build $\mathbf{V} = \mathbf{A}\sigma_a^2 + \mathbf{I}\sigma_e^2$, calculate the multivariate normal log-likelihood, and minimize the negative log-likelihood.

```{r}
library(mvtnorm)

set.seed(2026)

# the likelihood for the variance components.
# uniform above 0 and zero if negative
varlik <- function(x) {

  res <- 0
  if (x > 0) res = 1

  return(log(res))

}

# a small relationship matrix for a teaching example
n_animals <- 20
A <- diag(n_animals)
A[1, 2] <- A[2, 1] <- 0.5
A[3, 4] <- A[4, 3] <- 0.5
A[5, 6] <- A[6, 5] <- 0.25

# design matrix for the intercept
X <- matrix(1, n_animals, 1)

# true values used to simulate the data
b <- 30
var_a <- 4
var_e <- 9

# phenotype
a <- as.numeric(rmvnorm(1, mean = rep(0, n_animals), sigma = A * var_a))
y <- as.numeric(X %*% b + a + rnorm(n_animals, sd = sqrt(var_e)))

# likelihood function = L(y;.) = prod(PDF(y, E(y) = Xb, var(y) = A var_a + I var_e))
lik <- function(par, y, X, A) {

  # likelihood for the variance components
  L2 <- varlik(par[1])
  L3 <- varlik(par[2])

  # if any variance is outside the allowed range, stop here
  if (!all(is.finite(c(L2, L3)))) return(Inf)

  # it has to be this awkward because 'optim' only wants one vector of parameters
  b <- matrix(c(par[3:(2 + ncol(X))]), ncol = 1)

  # this is A var_a
  G <- A * par[2]

  # this is I var_e
  R <- diag(length(y)) * par[1]

  # this is var(y)
  V <- G + R

  # E(y)
  mu <- X %*% b

  # optim minimizes - therefore we need the negative likelihood
  # Instead of taking the products over the likelihoods we take
  # the sum over log(L) for numerical stability
  L1 <- tryCatch(
    dmvnorm(x = y, mean = as.numeric(mu), sigma = V, log = TRUE),
    error = function(e) -Inf
  )

  if (!all(is.finite(L1))) return(Inf)

  return(-sum(L1 + L2 + L3))

}


# This is only necessary because 'optim' wants one parameter vector
getp <- function(X, b = 0.1, var_e = 1, var_a = 1) {

  p <- c(var_e, var_a, rep(b, ncol(X)))
  names(p) <- c("var_e", "var_a", paste("b", 0:(ncol(X) - 1), sep = ""))

  p

}


# Get the Maximum Likelihood estimates for our unknowns
methods = c("Nelder-Mead", "BFGS", "CG", "L-BFGS-B", "SANN", "Brent")
ML <- optim(getp(X, b = mean(y)), lik, X = X, A = A, y = y, method = methods[1])

# compare ML results to the true simulation values
out <- rbind(
  ML = ML$par,
  true = c(var_e, var_a, b)
)

out
```

## Exercises

1. Change the starting values in `p` for the normal mean example. Does the optimum change?
2. Estimate $\mu$ and $\sigma$ for a smaller sample size, for example `n = 20`.
3. Change the true value of `rho` in the correlation example and rerun the analysis.
4. In the animal model example, change the relationship matrix `A`. What happens to the variance component estimates?
5. Explain why each likelihood function returns `-sum(L)` rather than `sum(L)`.
