---
title: "Multivariate Clustering"
subtitle: "PCA, k-means, and multivariate mixture models"
format:
  html:
    code-fold: false
---

<div class="view-switch"><a href="../lectures/multivariate-clustering.html">Open slides</a></div>

Multivariate clustering uses several features at once. In animal breeding, a common example is population structure from genotype data: animals may form groups because of breed, line, family, or selection history.

```{r}
#| include: false
library(ggplot2)
library(knitr)

theme_set(theme_minimal(base_size = 13))
```

## Multivariate Mixture Model

For an observation vector $\mathbf{x}_i \in \mathbb{R}^p$, a multivariate mixture model is

$$
p(\mathbf{x}_i)
=
\sum_{k=1}^K \pi_k f_k(\mathbf{x}_i \mid \boldsymbol{\theta}_k).
$$

With multivariate normal components,

$$
f_k(\mathbf{x}_i \mid \boldsymbol{\theta}_k)
=
N_p(\mathbf{x}_i \mid \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k).
$$

The squared Mahalanobis distance to component $k$ is

$$
d_{ik}^2
=
(\mathbf{x}_i - \boldsymbol{\mu}_k)'
\boldsymbol{\Sigma}_k^{-1}
(\mathbf{x}_i - \boldsymbol{\mu}_k).
$$

For two features, equal-density contours are ellipses.

## Simulate Population Structure

We simulate marker genotypes for three related but distinct lines. The labels are kept only for checking the clustering result.

```{r}
simulate_structured_genotypes <- function(n_per_line = c(70, 70, 70), n_markers = 450, seed = 2026) {
  set.seed(seed)
  base_frequency <- runif(n_markers, min = 0.15, max = 0.45)
  line_shift <- c(-0.06, 0.00, 0.07)

  marker_list <- lapply(seq_along(n_per_line), function(line) {
    p <- base_frequency + line_shift[line] + rnorm(n_markers, mean = 0, sd = 0.015)
    p <- pmin(pmax(p, 0.02), 0.98)
    sapply(p, function(marker_p) rbinom(n_per_line[line], size = 2, prob = marker_p))
  })

  list(
    M = do.call(rbind, marker_list),
    line = factor(rep(paste("line", LETTERS[seq_along(n_per_line)]), n_per_line))
  )
}

geno <- simulate_structured_genotypes()

dim(geno$M)
table(geno$line)
```

## PCA As Feature Compression

Principal component analysis summarizes many correlated marker columns into orthogonal axes:

$$
\mathbf{X}
=
\mathbf{U}\mathbf{D}\mathbf{V}'.
$$

The principal component scores are in $\mathbf{U}\mathbf{D}$.

```{r}
pca <- prcomp(geno$M, center = TRUE, scale. = TRUE)

pc_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  PC3 = pca$x[, 3],
  line = geno$line
)

variance_explained <- pca$sdev^2 / sum(pca$sdev^2)

kable(
  data.frame(
    component = paste0("PC", 1:6),
    variance_explained = variance_explained[1:6],
    cumulative = cumsum(variance_explained)[1:6]
  ),
  digits = 3
)
```

```{r}
ggplot(pc_data, aes(PC1, PC2, color = line)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("#0c6773", "#c46a2b", "#7a6f42")) +
  labs(x = "PC1", y = "PC2", color = "Known line")
```

## K-Means Clustering

K-means assigns observations to the closest cluster center by minimizing within-cluster squared distances:

$$
\sum_{k=1}^K \sum_{i:z_i=k}
\left\lVert \mathbf{x}_i - \boldsymbol{\mu}_k \right\rVert^2.
$$

```{r}
set.seed(2026)

km <- kmeans(pc_data[, c("PC1", "PC2", "PC3")], centers = 3, nstart = 50)
pc_data$kmeans_cluster <- factor(km$cluster)

kable(table(cluster = pc_data$kmeans_cluster, line = pc_data$line))
```

```{r}
ggplot(pc_data, aes(PC1, PC2, color = kmeans_cluster)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("#0c6773", "#c46a2b", "#7a6f42")) +
  labs(x = "PC1", y = "PC2", color = "K-means")
```

## Multivariate Normal Mixture

A multivariate normal mixture replaces hard nearest-center assignment with probabilistic assignment.

```{r}
dmvn_density <- function(X, mean, Sigma) {
  X <- as.matrix(X)
  p <- ncol(X)
  Sigma <- Sigma + diag(1e-6, p)
  centered <- sweep(X, 2, mean, "-")
  Sigma_inv <- solve(Sigma)
  quadratic <- rowSums((centered %*% Sigma_inv) * centered)
  log_det <- as.numeric(determinant(Sigma, logarithm = TRUE)$modulus)
  exp(-0.5 * (p * log(2 * pi) + log_det + quadratic))
}

fit_mvn_mixture <- function(X, k = 3, max_iter = 150, tolerance = 1e-6) {
  X <- as.matrix(X)
  n <- nrow(X)
  p <- ncol(X)

  km_init <- kmeans(X, centers = k, nstart = 30)
  pi <- as.numeric(table(factor(km_init$cluster, levels = seq_len(k)))) / n
  mu <- do.call(rbind, lapply(seq_len(k), function(j) colMeans(X[km_init$cluster == j, , drop = FALSE])))
  Sigma <- lapply(seq_len(k), function(j) {
    cov(X[km_init$cluster == j, , drop = FALSE]) + diag(1e-6, p)
  })

  loglik <- rep(NA_real_, max_iter)

  for (iteration in seq_len(max_iter)) {
    density <- vapply(seq_len(k), function(j) pi[j] * dmvn_density(X, mu[j, ], Sigma[[j]]), numeric(n))
    responsibility <- density / rowSums(density)

    nk <- colSums(responsibility)
    pi <- nk / n

    for (j in seq_len(k)) {
      mu[j, ] <- colSums(X * responsibility[, j]) / nk[j]
      centered <- sweep(X, 2, mu[j, ], "-")
      Sigma[[j]] <- crossprod(centered * sqrt(responsibility[, j])) / nk[j] + diag(1e-6, p)
    }

    density_updated <- vapply(seq_len(k), function(j) pi[j] * dmvn_density(X, mu[j, ], Sigma[[j]]), numeric(n))
    loglik[iteration] <- sum(log(rowSums(density_updated)))

    if (iteration > 1 && abs(loglik[iteration] - loglik[iteration - 1]) < tolerance) {
      loglik <- loglik[seq_len(iteration)]
      break
    }
  }

  density <- vapply(seq_len(k), function(j) pi[j] * dmvn_density(X, mu[j, ], Sigma[[j]]), numeric(n))
  responsibility <- density / rowSums(density)

  list(
    pi = pi,
    mu = mu,
    Sigma = Sigma,
    responsibility = responsibility,
    cluster = max.col(responsibility),
    loglik = loglik[length(loglik)]
  )
}

X2 <- as.matrix(pc_data[, c("PC1", "PC2")])
set.seed(2026)
mvn_fit <- fit_mvn_mixture(X2, k = 3)

pc_data$mvn_cluster <- factor(mvn_fit$cluster)
pc_data$mvn_uncertainty <- 1 - apply(mvn_fit$responsibility, 1, max)

kable(table(cluster = pc_data$mvn_cluster, line = pc_data$line))
```

## Ellipses

An observation is inside the 95% ellipse of component $k$ when

$$
d_{ik}^2 \leq \chi^2_{p, 0.95}.
$$

```{r}
ellipse_points <- function(center, Sigma, level = 0.90, n = 200) {
  angle <- seq(0, 2 * pi, length.out = n)
  circle <- cbind(cos(angle), sin(angle)) * sqrt(qchisq(level, df = 2))
  points <- circle %*% chol(Sigma + diag(1e-6, 2))
  data.frame(
    PC1 = points[, 1] + center[1],
    PC2 = points[, 2] + center[2]
  )
}

ellipse_data <- do.call(
  rbind,
  lapply(seq_len(3), function(j) {
    out <- ellipse_points(mvn_fit$mu[j, ], mvn_fit$Sigma[[j]], level = 0.90)
    out$cluster <- factor(j)
    out
  })
)

ggplot(pc_data, aes(PC1, PC2, color = mvn_cluster)) +
  geom_point(aes(alpha = 1 - mvn_uncertainty)) +
  geom_path(data = ellipse_data, aes(PC1, PC2, color = cluster), linewidth = 1) +
  scale_color_manual(values = c("#0c6773", "#c46a2b", "#7a6f42")) +
  scale_alpha(range = c(0.35, 0.9), guide = "none") +
  labs(x = "PC1", y = "PC2", color = "Mixture")
```

## Choosing K

For a full-covariance multivariate normal mixture in $p$ dimensions, the number of free parameters is

$$
q
=
(K - 1)
+ Kp
+ K\frac{p(p + 1)}{2}.
$$

```{r}
bic_mvn <- do.call(
  rbind,
  lapply(1:5, function(k) {
    set.seed(100 + k)
    fit <- fit_mvn_mixture(X2, k = k)
    p <- ncol(X2)
    q <- (k - 1) + k * p + k * p * (p + 1) / 2
    data.frame(
      components = k,
      loglik = fit$loglik,
      bic = -2 * fit$loglik + q * log(nrow(X2))
    )
  })
)

kable(bic_mvn, digits = 2)
```

```{r}
ggplot(bic_mvn, aes(components, bic)) +
  geom_line(color = "#0c6773") +
  geom_point(size = 2.4, color = "#c46a2b") +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Number of mixture components", y = "BIC")
```

## Exercises

1. Cluster using only `PC1` and `PC2`. Compare with using `PC1`, `PC2`, and `PC3`.
2. Increase `n_markers`. Does the PCA separation become clearer?
3. Change the line shifts in the genotype simulation and rerun the clustering.
4. Compare k-means and mixture assignments for uncertain observations.
5. Explain why PCA is often useful before clustering marker data.
